---
title: "Mitochondrial drift analysis (blood)"
author: "Mike Spencer Chapman"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
cran_packages=c("devtools","ape","stringr","dplyr","tidyr","ggplot2","gridExtra","gganimate","ggridges","abc","gifski")

for(package in cran_packages){
  if(!require(package, character.only=T,quietly = T, warn.conflicts = F)){
    install.packages(as.character(package),repos = "http://cran.us.r-project.org")
    library(package, character.only=T,quietly = T, warn.conflicts = F)
  }
}

#Install & load github packages
if(!require("dndscv", character.only=T,quietly = T, warn.conflicts = F)){
  devtools::install_github("im3sanger/dndscv")
  library("dndscv",character.only=T,quietly = T, warn.conflicts = F)
}
```

Read in the blood mitochondrial mutation data for the analysis. 

```{r setup2,include=T}
source(here::here("config.R")) #sets root_dir, genomeFile, project paths and my_theme
plots_dir=notebook_plots_dir("Mitochondrial_drift_analysis")
figures_dir=notebook_plots_dir("Mitochondrial_drift_analysis") #see save_plots in config.R
source(paste0(root_dir,"/data/mito_mutations_blood_functions.R"))
mito_data_file=paste0(root_dir,"/data/mito_data.Rds")
ref_file=paste0(root_dir,"/data/Samples_metadata_ref.csv")

mito_data<-readRDS(mito_data_file)
ref_df=read.csv(ref_file)%>%filter(Dataset!="Lymphocyte")
Individual_cols=RColorBrewer::brewer.pal(12,"Paired")
names(Individual_cols)<-ref_df$Sample[order(ref_df$Age)]

#Read in the mutation signature by VAF and sample data
exposures<-read.csv(paste0(root_dir,"/data/mutational_signatures/exposures.csv"))
components<-read.csv(paste0(root_dir,"/data/mutational_signatures/components.csv"))
mutation_profiles_mat=read.table(paste0(root_dir,"/data/mutational_signatures/trinuc_mut_mat.txt"))
sig_ref<-readRDS(file=paste0(root_dir,"/data/mutational_signatures/sig_ref_file.Rds"))

#Clean up the row and column names
sub_vec = c("C>A","C>G","C>T","T>A","T>C","T>G"); ctx_vec = paste(rep(c("A","C","G","T"),each=4),rep(c("A","C","G","T"),times=4),sep="-")
full_vec = paste(rep(sub_vec,each=16),rep(ctx_vec,times=6),sep=",")
colnames(components)<-c(paste(full_vec,"H",sep="_"),paste(full_vec,"L",sep="_"))
rownames(components)<-paste0("N",0:(nrow(components)-1))
rownames(exposures)<-paste0("N",0:(nrow(exposures)-1))
colnames(exposures)<-gsub("^X","",colnames(exposures))
colnames(exposures)<-gsub("\\.pcw"," pcw",colnames(exposures))

#Set the basic plotting theme for ggplot2
my_theme<-theme(text = element_text(family="Helvetica"),
                axis.text = element_text(size = 5),
                axis.title = element_text(size=7),
                legend.text = element_text(size=5),
                legend.title = element_text(size=7),
                strip.text = element_text(size=7),
                legend.spacing = unit(1,"mm"),
                legend.key.size= unit(5,"mm"))+
  theme(legend.key.height=unit(3,"mm"),
        legend.title = element_text(size=8))

my_rmd_theme<-theme(text = element_text(family="Helvetica"),
                axis.text = element_text(size = 8),
                axis.title = element_text(size=10),
                legend.text = element_text(size=8),
                legend.title = element_text(size=10),
                strip.text = element_text(size=11),
                legend.spacing = unit(1.2,"mm"),
                legend.key.size= unit(7,"mm"))+
  theme(legend.key.height=unit(4,"mm"),
        legend.title = element_text(size=10))
```


```{r}
VAF_groups_original=unique(sapply(stringr::str_split(colnames(exposures),pattern="_"),function(vec) paste(vec[2:3],collapse="_")))
new_VAF_groups=c("<0.1%","0.1-0.2%","0.2-0.4%","0.4-0.8%","0.8-1.6%","1.6-3.1%","3.1-6.2%","6.2-12.5%","12.5-25%","25-50%",">50%")
names(new_VAF_groups)<-VAF_groups_original

#Define the VAF 'bins' that will be used for comparing VAF distributions
VAF_groups=data.frame(
  labels=new_VAF_groups,
  lower_limit=c(0,2^(-10:-1)),
  upper_limit=2^(-10:0)
)

VAF_groups
```


```{r}
mut_numbers=data.frame(SampleID=rownames(mutation_profiles_mat),nmuts=rowSums(mutation_profiles_mat))

#Fill in the rare bins with too low mutation numbers - assume all are N1, as these are high VAF mutations
# (in samples with adequate mutations for extraction, high VAF mutations are all N1)
excluded_cats<-rownames(mutation_profiles_mat)[!rownames(mutation_profiles_mat)%in%colnames(exposures)]
all_N1=c(0,1,rep(0,nrow(exposures)-2))
excluded_mat<-matrix(rep(all_N1,times=length(excluded_cats)),nrow=length(all_N1),dimnames = list(rownames(exposures),excluded_cats))

real_muts_dist_df<-t(cbind(exposures,excluded_mat))%>%
  as.data.frame()%>%
  tibble::rownames_to_column(var="SampleID")%>%
  left_join(mut_numbers)%>%
  mutate(exp_ID=stringr::str_split(SampleID,pattern="_",simplify=T)[,1])%>%
  mutate(VAF_range=sapply(stringr::str_split(SampleID,pattern="_"),function(vec) paste(vec[2:3],collapse="_")))%>%
  mutate(VAF_range=factor(new_VAF_groups[VAF_range],levels=new_VAF_groups))%>%
  dplyr::select(-SampleID)%>%
  gather(-VAF_range,-exp_ID,-nmuts,key="Signature",value="Exposure")%>%
  mutate(abs_muts=Exposure*nmuts)%>%
  mutate(Signature=factor(Signature,levels=c(rownames(exposures)[nrow(exposures):3],"N0","N1")))%>%
  dplyr::filter(Signature=="N1")%>%
  mutate(exp_ID=gsub("8 pcw","8pcw",exp_ID))

head(real_muts_dist_df)
```


```{r fig.width=8,fig.height=4}
real_mut_distribution_plot<-real_muts_dist_df%>%
  ggplot(aes(x=VAF_range,y=abs_muts))+
  geom_bar(stat="identity",fill="#FDBF6F",col="black",linewidth=0.25)+
  geom_line(aes(group=exp_ID))+
  theme_bw()+
  my_theme+
  theme(axis.text.x = element_text(size=5,angle=90),strip.text.x = element_text(size=6))+
  #facet_grid(rows=vars(factor(exp_ID,levels=ref_df$Sample[order(ref_df$Age)])),scales="free")+
  facet_wrap(~factor(exp_ID,levels=ref_df$Sample[order(ref_df$Age)]),scales="fixed",dir="v",ncol=6)+
  labs(x="VAF range",y="Number of N1 mutations detected")

real_mut_distribution_plot+my_rmd_theme+theme(strip.text.x = element_text(size=9),axis.text.x = element_text(size=6,angle=90))
```

Normalize mutation numbers by the number of samples included.

```{r fig.width=8,fig.height=4}
nsamp_df<-Map(list=mito_data,Exp_ID=names(mito_data),function(list,Exp_ID) {
  data.frame(exp_ID=Exp_ID,n_samp=length(colnames(list$matrices$SW)))
})%>%dplyr::bind_rows()%>%
  mutate(exp_ID=gsub("8 pcw","8pcw",exp_ID))

real_mut_distribution_plot_normalized<-left_join(real_muts_dist_df,nsamp_df)%>%
  mutate(muts_per_samp=abs_muts/n_samp)%>%
  ggplot(aes(x=VAF_range,y=muts_per_samp))+
  geom_bar(stat="identity",fill="#FDBF6F",col="black",linewidth=0.25)+
  geom_line(aes(group=exp_ID))+
  theme_bw()+
  my_theme+
  theme(axis.text.x = element_text(size=5,angle=90),strip.text.x = element_text(size=6))+
  #facet_grid(rows=vars(factor(exp_ID,levels=ref_df$Sample[order(ref_df$Age)])),scales="free")+
  facet_wrap(~factor(exp_ID,levels=ref_df$Sample[order(ref_df$Age)]),scales="fixed",dir="v",ncol=6)+
  labs(x="VAF range",y="Number of N1 mutations detected per sample")

real_mut_distribution_plot_normalized+my_rmd_theme+theme(strip.text.x = element_text(size=9),axis.text.x = element_text(size=6,angle=90))
save_plot(filename=paste0(figures_dir,"real_mut_distribution_plot.pdf"),real_mut_distribution_plot_normalized,width=7,height=2.5)
```


```{r}
sumstats.data<-left_join(real_muts_dist_df,nsamp_df)%>%
  mutate(muts_per_samp=abs_muts/n_samp)%>%
  dplyr::select(exp_ID,VAF_range,muts_per_samp)%>%
  pivot_wider(names_from="VAF_range",values_from="muts_per_samp")%>%
  replace(is.na(.), 0)

head(sumstats.data)
```

### VISUALIZE WRIGHT-FISHER MODELS OF DRIFT
This code is to plot the shifting distribution of VAFs over time - single run of simulation but storing information after each generation

```{r}
muts_per_mitochondria_per_generation=5e-4
print(muts_per_mitochondria_per_generation)
mito_copy_number=600 #This is the
number_of_cells_in_simulation=1000 #Need enough 'cells' to have adequate mutation numbers to define the distribution
mutation_introductions_per_generation=muts_per_mitochondria_per_generation*number_of_cells_in_simulation*mito_copy_number

```

Record the distribution of VAFs every 10 generations, though the generation of mutation acquisition is recorded exactly.

```{r}
gens_to_include=seq(10,1500,10)
gens_record=vector(mode="list",length = length(gens_to_include))
vaf_by_gen_list=vector(mode="list")
last_gen=0
curr_vafs=c()
for(ngen in gens_to_include) {
  #print(ngen)
  if(length(vaf_by_gen_list)>0) {
    drifted_old_muts=lapply(vaf_by_gen_list,function(gen_vafs) {
      new_gen_vafs<-sapply(gen_vafs,function(vaf) fisher_wright_drift(vaf,population_size = mito_copy_number,1,ngen-last_gen))
      return(new_gen_vafs[new_gen_vafs>0])
    })
  } else {
    drifted_old_muts<-NULL
  }
  
  new_muts_gen_list=vector(mode="list",length=ngen-last_gen)
  names(new_muts_gen_list)<-(last_gen+1):ngen
  for(j in 1:(ngen-last_gen)){
    gen_mut_vafs<-vector(length=mutation_introductions_per_generation)
    for(i in 1:mutation_introductions_per_generation) {
      this_mut_final_vaf<- fisher_wright_drift(1/mito_copy_number,population_size = mito_copy_number,1,j)
      gen_mut_vafs[i]<-this_mut_final_vaf
    }
    new_muts_gen_list[[j]]<-gen_mut_vafs[gen_mut_vafs>0]
  }
  vaf_by_gen_list<-c(drifted_old_muts,new_muts_gen_list)
  gen_summary=Map(vafs=vaf_by_gen_list,gen=names(vaf_by_gen_list),function(vafs,gen) if(length(vafs)>0){data.frame(gen=gen,vaf=vafs)}else {NULL})%>%dplyr::bind_rows()%>%mutate(total_gens=ngen)
  gens_record[[which(gens_to_include==ngen)]]<-gen_summary
  last_gen<-ngen
}
```

Here, 7.5 is the 'haploid sequencing coverage' in this simulated experiment.
This approximates the ~15X diploid coverage of most of the experiments.

```{r}
haploid_coverage=7.5
gens_record<-dplyr::bind_rows(gens_record)
gens_record$observed_vaf=sapply(gens_record$vaf,function(vaf) rbinom(n=1,size=(haploid_coverage*mito_copy_number),prob=vaf)/(haploid_coverage*mito_copy_number))

gens_record$VAF_group=sapply(1:nrow(gens_record),function(i) {
  if(gens_record$observed_vaf[i]==0) {
    return("Absent")
  } else {
    return(VAF_groups$labels[VAF_groups$lower_limit<gens_record$observed_vaf[i] & VAF_groups$upper_limit>=gens_record$observed_vaf[i]])
  }
})
```

Visualize the distribution at a selection of "generations" to get a sense of the evolution of the distribution through time

```{r fig.width=7,fig.height=2.5}
modelled_VAF_dist<-gens_record%>%
  filter(total_gens%in%c(100,200,400,800,1200))%>%
  mutate(total_gens=factor(paste(total_gens,"generations"),levels=paste(c(100,200,400,800,1200),"generations")))%>%
  mutate(VAF_group=factor(VAF_group,levels=new_VAF_groups),gen=factor(gen,levels=1:1200))%>%
  ggplot(aes(x=VAF_group,fill=gen))+
  geom_bar(stat="count")+
  facet_wrap(~total_gens,ncol=5)+
  theme_bw()+
  theme(legend.position = "none",axis.text.x = element_text(size=5,angle=90),strip.text.x = element_text(size=6))+
  scale_fill_manual(values = colorRampPalette(RColorBrewer::brewer.pal(8,"Spectral"))(1200))+
  labs(x="VAF group",y="Count")+
  my_theme

modelled_VAF_dist+my_rmd_theme+theme(strip.text.x = element_text(size=9),axis.text.x = element_text(size=6,angle=90))
save_plot(filename=paste0(figures_dir,"Modelled_VAF_dist.pdf"),modelled_VAF_dist,width = 6,height=2)
```

How many 'high VAF' (>50%) mutaitons are there depending on the generation?

```{r fig.width=3,fig.height=2.5}
gens_record%>%
  filter(VAF_group==">50%")%>%
  #mutate(total_gens=factor(total_gens,levels=seq(10,1500,10)))%>%
  dplyr::group_by(total_gens)%>%
  summarise(n=n(),.groups = "drop_last")%>%
  ggplot(aes(x=total_gens,y=n))+
  scale_x_continuous(limits=c(0,1500))+
  geom_point(size=0.3)+
  geom_line(linewidth=0.3)+
  theme_classic()+
  my_rmd_theme
```

One notable feature is that these simulations suggest that high VAF mutations were acquired early in life.
Conversely, low VAF mutations were invariably acquired very recently.

```{r fig.width=5,fig.height=4}
acquisition_time_by_VAF<-gens_record%>%
  mutate(VAF_group=factor(VAF_group,levels=VAF_groups$labels))%>%
  filter(total_gens==1250)%>%
  ggplot(aes(x=as.numeric(gen),y=VAF_group))+
  ggridges::geom_density_ridges(panel_scaling = F)+
  theme_classic()+
  my_theme+
  theme(axis.text.x=element_text(angle=90),strip.text.x = element_text(size=7,margin = unit(c(1,0,1,0),"mm")))+
  labs(x="Time of mutation acquisition\n(WF Generations)",y="Density")

acquisition_time_by_VAF+my_rmd_theme
save_plot(filename = paste0(figures_dir,"Acquisition_time_by_VAF.pdf"),acquisition_time_by_VAF,width=7,height=2.5)
```

Plot same data but as a ridges plot.
```{r fig.width=5,fig.height=4}
acquisition_time_by_VAF_ridges<-gens_record%>%
  mutate(VAF_group=factor(VAF_group,levels=VAF_groups$labels))%>%
  filter(total_gens==1250)%>%
  ggplot(aes(x=as.numeric(gen),y=VAF_group,fill = factor(stat(quantile),levels=1:4))) +
  stat_density_ridges(geom = "density_ridges_gradient", calc_ecdf = TRUE,quantiles = 4, quantile_lines = TRUE,scale=4) +
  scale_fill_viridis_d(name = "Quartiles")+
  scale_x_continuous(limits=c(0,1350),breaks=seq(0,1250,250))+
  theme_classic()+
  my_theme+
  labs(x="Time of mutation acquisition\n(WF Generations)",y="VAF level")

acquisition_time_by_VAF_ridges+my_rmd_theme
save_plot(filename = paste0(figures_dir,"acquisition_time_by_VAF_ridges.pdf"),acquisition_time_by_VAF_ridges,width=3.3,height=2.5)
```
### Animation of the Wright-Fisher model

Can also be informative to visualize the evolving distribution over time as an animation.

```{r fig.width=5,fig.height=4}
gens_record_summary<-gens_record%>%
  group_by(total_gens,gen,VAF_group)%>%
  summarise(n=n(),.groups = "drop_last")
gens_record_summary$VAF_group=factor(gens_record_summary$VAF_group,levels=VAF_groups$labels)
gens_record_summary$gen=as.numeric(gens_record_summary$gen)

drift_with_age_animation<-gens_record_summary%>%
  filter(!is.na(VAF_group))%>%
  arrange(gen)%>%
  ggplot(aes(x=VAF_group,y=n,fill=gen))+
  geom_bar(stat="identity")+
  transition_manual(frames=total_gens)+
  theme_bw()+
  my_theme+
  theme(title = element_text(size=10),legend.position = "none",axis.text.x = element_text(angle=90,size=10),axis.title.x=element_text(size=8),axis.title.y=element_text(size=10))+
  scale_fill_gradientn(colours = RColorBrewer::brewer.pal(8,"Spectral"))+
  labs(title='{current_frame} generations',x="Observed VAF",y="Count")

drift_with_age_animation+my_rmd_theme
anim <- gganimate::animate(drift_with_age_animation,renderer = gifski_renderer())
save_anim(filename=paste0(rebuttal_figs_dir,"driftwithage.gif"),animation = anim)
```

Plot the colour scale for these figures.
```{r fig.width=1,fig.height=2.5}
max_gen=1500
lut=colorRampPalette(RColorBrewer::brewer.pal(8,"Spectral"))(max_gen)
scale = (length(lut)-1)
save_pdf(paste0(rebuttal_figs_dir,"mito_muts_by_gen_scale.pdf"),height=6,width = 3)
plot(c(0,10), c(0,1), type='n', bty='n', xaxt='n', xlab='', yaxt='n', ylab='', main="")
axis(side=2,pos=1,las=1,at=seq(0,max_gen,200)/scale,labels=seq(0,max_gen,200))
for (i in 1:(length(lut)-1)) {
  y = (i-1)/scale
  rect(1,y,2,y+1/scale, col=lut[i], border=NA)
}
dev.off()
```

### PERFORM THE ABC - with all mutations.

This section uses summary statistics collected from many simulations and compares the results to those from the data stastics for each individual (as collected above).
The actual simulations are run in the separate script "Mito_VAF_distribution_simulations.R" and saved in an Rds object.

```{r message=F,results="hide"}
all_sumstats=readRDS(file=paste0(root_dir,"/data/Drift_ABC_VAF_distribution/VAF_distribution_ABC_simulation_sumstats_combined.Rds"))

new_VAF_groups=c("<0.1%","0.1-0.2%","0.2-0.4%","0.4-0.8%","0.8-1.6%","1.6-3.1%","3.1-6.2%","6.2-12.5%","12.5-25%","25-50%",">50%")

#Exclude the lowest VAF categories from abc that may be unreliably captured by sequencing
VAF_categories_to_include_in_ABC=new_VAF_groups[3:11]

mtDNA_CN=600
abc_res_df<-lapply(1:nrow(sumstats.data),function(i) {
  Exp_ID=sumstats.data[i,1]
  res<-abc(target = as.numeric(sumstats.data[i,VAF_categories_to_include_in_ABC]),
           param = all_sumstats[,1:2],
           sumstat = all_sumstats[,VAF_categories_to_include_in_ABC],
           tol = 0.02,
           transf = c("log","log"),
           method = "neuralnet")
  stats=apply(res$adj.values,2,quantile,c(0.025,0.5,0.975))
  stats=as.data.frame(stats)%>%tibble::rownames_to_column(var="quantile")%>%pivot_wider(names_from = "quantile",values_from=c("muts_per_mitochondria_per_generation","total_generations"))
  return(cbind(data.frame(exp_ID=Exp_ID),stats))
})%>%dplyr::bind_rows()%>%
  left_join(ref_df,by=c("exp_ID"="Sample"))
```


```{r fig.width=5,fig.height=4}
abc_res_plot_mut_rate<-abc_res_df%>%
  filter(Age>=0)%>%
  ggplot(aes(x=forcats::fct_reorder(exp_ID,Age),y=`muts_per_mitochondria_per_generation_50%`,
             ymin=`muts_per_mitochondria_per_generation_2.5%`,
             ymax=`muts_per_mitochondria_per_generation_97.5%`))+
  geom_smooth(method="lm",col="black",linewidth=0.5)+
  geom_point(alpha=0.75,size=0.5)+
  geom_errorbar(width=0.3,alpha=0.5)+
  theme_classic()+
  scale_y_continuous(limits=c(1e-4,8e-4),breaks=seq(2e-4,1e-3,2e-4))+
  labs(x="Individual",y="Mutations per mitochondria\nper generation")+
  my_theme+
  coord_flip()

abc_res_plot_mut_rate+my_rmd_theme
save_plot(filename=paste0(figures_dir,"abc_res_plot_mut_rate.pdf"),abc_res_plot_mut_rate,width=1.75,height=2)
```

Now do linear mixed effects modelling using all values of the posterior distribution and using individual as a random effect.
First, run an abc on each individual, one at a time, using neural network regression to improve the posterior distributions.
Each individual is a row of the sumstats.data matrix.
Then bind the posterior distribution results from each abc and add in the related metadata for each individual (primarily age).
However, do not include the foetal samples as the distributions at this very early time point are likely dominated by heteroplasmic oocyte mutations which are not explicitly modelled in the simulations.

```{r message=F,warning=F,results="hide"}
abc_res_df<-lapply(1:nrow(sumstats.data),function(i) {
  Exp_ID=sumstats.data[i,1]
  res<-abc(target = as.numeric(sumstats.data[i,VAF_categories_to_include_in_ABC]),
           param = all_sumstats[,1:2],
           sumstat = all_sumstats[,VAF_categories_to_include_in_ABC],
           tol = 0.05,
           transf = c("log","log"),
           method = "neuralnet")
  
  return(as.data.frame(res$adj.values)%>%mutate(exp_ID=Exp_ID$exp_ID))
})%>%dplyr::bind_rows()%>%
  left_join(ref_df%>%dplyr::select(Sample,Age),by=c("exp_ID"="Sample"))%>%
  dplyr::filter(!exp_ID%in%c("8pcw","18pcw"))
```

Then perform the LMER

```{r}
lme.gens_by_age<-lme4::lmer(total_generations~Age+(1|exp_ID),data=abc_res_df)
summary(lme.gens_by_age)
mtDNA_CN*365/lme.gens_by_age@beta[2]
mtDNA_CN*365/confint(lme.gens_by_age)["Age",]
```

Plot these results.
```{r fig.width=5,fig.height=4}
abc_res_plot_total_generations<-abc_res_df%>%
  dplyr::filter(!exp_ID%in%c("8pcw","18pcw"))%>%
  mutate(exp_ID=factor(exp_ID,levels=ref_df$Sample[order(ref_df$Age)]))%>%
  ggplot(aes(x=Age,y=total_generations))+
  geom_point(aes(col=exp_ID),alpha=0.05,size=0.25)+
  scale_y_continuous(limits=c(0,1700))+
  geom_abline(slope=lme.gens_by_age@beta[2],intercept = lme.gens_by_age@beta[1],linetype=1)+
  scale_color_manual(values=Individual_cols[-c(1:2)])+
  theme_classic()+
  guides(colour=guide_legend(override.aes = list(alpha=1)))+
  labs(x="Age",y="Total WF generations\n(Posterior distribution from ABC)",col="")+
  my_theme+
  theme(legend.box.spacing = unit(0,"mm"),
        legend.key.size = unit(0.5,"mm"),
        legend.spacing = unit(0,"mm"),
        legend.box.margin = margin(c(0,0,0,0)),
        legend.text = element_text(margin = margin(t=0)))

abc_res_plot_total_generations+my_rmd_theme
save_plot(filename=paste0(figures_dir,"abc_res_plot_total_generations.pdf"),abc_res_plot_total_generations,width=2,height=2)
```

## Now a version of the ABC with only synonymous/ non-coding mutation

This was requested by reviewers to ensure that non-synonymous mutations are not introducing a bias in drift estimates due to their positive/ negative selection which may affect the heteroplasmy distributions.

Therefore, annotate all the mutations in the data, and retain only non-coding and synonymous variants.

```{r}
CN_correlating_muts<-readRDS(paste0(root_dir,"/data/CN_correlation.RDS"))
length(CN_correlating_muts)
exclude_muts=c("MT_16183_A_C") #This is a specific artefact primarily in the 8pcw foetus due to a germline SNP and consequent frequent PCR slippage

#Define the VAF bins to use for signature extraction
bins=2^c(-10:0)

#Generate tidy data frame of samples, mutations and vafs
mutCN_cutoff=25 #If the mitochondrial copy number is over 25, retain mutation even if is in the "CN correlating muts" list
df_tidy<-dplyr::bind_rows(Map(list=mito_data,exp_ID=names(mito_data),function(list,exp_ID) {
  if(is.null(list)){stop(return(NULL))}
  CN_correlating_mut_removal_mat=list$matrices$implied_mutCN>mutCN_cutoff|(matrix((!rownames(list$matrices$vaf)%in%CN_correlating_muts),ncol=1)%*%matrix(rep(1,ncol(list$matrices$vaf)),nrow=1))
  implied_mut_CN_tidy<-list$matrices$implied_mutCN%>%
    as.data.frame()%>%
    tibble::rownames_to_column(var="mut_ref")%>%
    dplyr::select(-global)%>%
    tidyr::gather(key="Sample",value="implied_mut_CN",-mut_ref)
  df_tidy<-(list$matrices$vaf*list$matrices$SW*(list$matrices$ML_Sig=="N1")*CN_correlating_mut_removal_mat)%>%
    as.data.frame()%>%
    tibble::rownames_to_column(var="mut_ref")%>%
    mutate(rho_val=list$rho_vals)%>%
    dplyr::select(-global)%>%
    tidyr::gather(key="Sample",value="vaf",-mut_ref,-rho_val)
  
  comb_tidy<-left_join(df_tidy,implied_mut_CN_tidy,by=c("Sample","mut_ref"))%>%
    dplyr::filter(!grepl("DEL|INS",mut_ref) & !mut_ref%in%exclude_muts)%>%
    dplyr::filter(vaf>0)%>%
    mutate(exp_ID=exp_ID)
  return(comb_tidy)
}))
```

Annotate the mutations using dndscv.
Excluding the missense/ nonsense/ stop loss mutations gets rid of almost half of mutations.

```{r message=F,warning=F}
mtref_rda_path=paste0(root_dir,"/data/mtref.rda")

#All mtDNA genes
target_genes <- c("MT-CYB", "MT-ND5", "MT-ND2", "MT-ND4", "MT-ND1", "MT-CO3", "MT-ATP6","MT-ND3", "MT-ATP8", "MT-ND4L", "MT-CO2", "MT-CO1","MT-ND5")

dndscv_input<-df_tidy%>%tidyr::separate(col=mut_ref,into = c("chr","pos","ref","alt"),sep = "_")%>%
  dplyr::select("sampleID"=Sample,chr,pos,ref,alt)

dndscv_output <- dndscv(dndscv_input,
                        gene_list=target_genes, 
                        refdb = mtref_rda_path,
                        numcode=2, #numcode = 2 specifies using the mitochondrial genetic code
                        max_coding_muts_per_sample = Inf,
                        max_muts_per_gene_per_sample = Inf)

mtDNA.annotated<-dndscv_output$annotmuts%>%
  tidyr::unite(col = "mut_ref",chr,pos,ref,mut,sep = "_")%>%
  dplyr::rename("Sample"=sampleID)

df_tidy_annotated<-left_join(df_tidy,mtDNA.annotated,by=c("Sample","mut_ref"))%>%
  tidyr::replace_na(replace=list(impact="Non-coding"))
table(df_tidy_annotated$impact)
df_tidy_syn_noncoding_only<-df_tidy_annotated%>%filter(impact=="Synonymous"|impact=="Non-coding")
```

Visualize the distribution of these.
Note that this is visually a simlar shape to the distributions from all the mutations, but with lower absolute mutations per cell.

```{r fig.width=8,fig.height=4}
syn_muts_distribution_df<-lapply(names(mito_data),function(this_exp_ID) {
  #print(this_exp_ID)
  exp_bins<-lapply(1:length(bins),function(i) {
    lower_limit=ifelse(i==1,0,bins[i-1])
    upper_limit=bins[i]
    #print(paste(lower_limit,"to",upper_limit))
    bin_exp_muts<-df_tidy_syn_noncoding_only%>%
      dplyr::filter(exp_ID==this_exp_ID & vaf>lower_limit & vaf<= upper_limit)%>%
      dplyr::distinct(exp_ID,mut_ref,.keep_all = T)%>%
      nrow()
    
    data.frame(exp_ID=this_exp_ID,VAF_range=VAF_groups$labels[i],Signature="N1",abs_muts=bin_exp_muts)
  })
  return(exp_bins)
})%>%dplyr::bind_rows()

#Bind together mutation sets & annotate the trinucleotide context
real_syn_mut_distribution_plot_normalized<-left_join(syn_muts_distribution_df,nsamp_df)%>%
  mutate(muts_per_samp=abs_muts/n_samp,VAF_range=factor(VAF_range,levels=new_VAF_groups))%>%
  ggplot(aes(x=VAF_range,y=muts_per_samp))+
  geom_bar(stat="identity",fill="#FDBF6F",col="black",linewidth=0.25)+
  geom_line(aes(group=exp_ID))+
  theme_bw()+
  my_theme+
  theme(axis.text.x = element_text(size=5,angle=90),strip.text.x = element_text(size=6))+
  #facet_grid(rows=vars(factor(exp_ID,levels=ref_df$Sample[order(ref_df$Age)])),scales="free")+
  facet_wrap(~factor(exp_ID,levels=ref_df$Sample[order(ref_df$Age)]),scales="fixed",dir="v",ncol=6)+
  labs(x="VAF range",y="Number of mutations not affecting coding sequence detected per sample")

real_syn_mut_distribution_plot_normalized+my_rmd_theme+theme(axis.text.x = element_text(size=7,angle=90),strip.text.x = element_text(size=10))

save_plot(filename=paste0(figures_dir,"synonymous_noncoding_distribution_plot.pdf"),real_mut_distribution_plot_normalized,width=7,height=2.5)
```

Create the matrix of summary statistics for the ABC.

```{r}
sumstats.syn.data<-left_join(syn_muts_distribution_df,nsamp_df,by="exp_ID")%>%
  mutate(muts_per_samp=abs_muts/n_samp)%>%
  dplyr::select(exp_ID,VAF_range,muts_per_samp)%>%
  pivot_wider(names_from="VAF_range",values_from="muts_per_samp")%>%
  replace(is.na(.), 0)
```


## Repeat the ABC with this mutation set (only non-coding/ synonymous)

This section uses summary statistics collected from many simulations and compares the results to those from the data stastics for each individual (as collected above).
The actual simulations are run in the separate script "Mito_VAF_distribution_simulations.R"

```{r message=F,warning=F,results="hide"}
all_sumstats=readRDS(file=paste0(root_dir,"/data/Drift_ABC_VAF_distribution/VAF_distribution_ABC_simulation_sumstats_combined.Rds"))

#Exclude the lowest VAF categories from abc that may be unreliably captured by sequencing
VAF_categories_to_include_in_ABC=new_VAF_groups[3:11]

mtDNA_CN=600
abc_syn_res_df<-lapply(1:nrow(sumstats.syn.data),function(i) {
  Exp_ID=sumstats.syn.data[i,1]
  res<-abc(target = as.numeric(sumstats.syn.data[i,VAF_categories_to_include_in_ABC]),
           param = all_sumstats[,1:2],
           sumstat = all_sumstats[,VAF_categories_to_include_in_ABC],
           tol = 0.02,
           transf = c("log","log"),
           method = "neuralnet")
  stats=apply(res$adj.values,2,quantile,c(0.025,0.5,0.975))
  stats=as.data.frame(stats)%>%tibble::rownames_to_column(var="quantile")%>%pivot_wider(names_from = "quantile",values_from=c("muts_per_mitochondria_per_generation","total_generations"))
  return(cbind(data.frame(exp_ID=Exp_ID),stats))
})%>%dplyr::bind_rows()%>%
  left_join(ref_df,by=c("exp_ID"="Sample"))

head(abc_syn_res_df)
```

The "synonymous/ non-coding" mutation rates look similar but ~half of the rate with the full mutation set (as expected given this is only ~50% of the mutations).

```{r fig.width=5,fig.height=4}
abc_synres_plot_mut_rate<-abc_syn_res_df%>%
  filter(Age>=0)%>%
  ggplot(aes(x=forcats::fct_reorder(exp_ID,Age),y=`muts_per_mitochondria_per_generation_50%`,
             ymin=`muts_per_mitochondria_per_generation_2.5%`,
             ymax=`muts_per_mitochondria_per_generation_97.5%`))+
  geom_smooth(method="lm",col="black",linewidth=0.5)+
  geom_point(alpha=0.75,size=0.5)+
  geom_errorbar(width=0.3,alpha=0.5)+
  theme_classic()+
  scale_y_continuous(limits=c(1e-4,8e-4),breaks=seq(2e-4,1e-3,2e-4))+
  labs(x="Individual",y="Mutations per mitochondria\nper generation")+
  my_theme+
  coord_flip()

abc_synres_plot_mut_rate+my_rmd_theme

save_plot(filename=paste0(figures_dir,"abc_synres_plot_mut_rate.pdf"),abc_synres_plot_mut_rate,width=1.75,height=2)
```

Now do linear mixed effects modelling using all values of the posterior distribution and using individual as a random effect.

```{r message=F,warning=F,results="hide"}
abc_syn_res_df<-lapply(1:nrow(sumstats.syn.data),function(i) {
  Exp_ID=sumstats.syn.data[i,1]
  res<-abc(target = as.numeric(sumstats.syn.data[i,VAF_categories_to_include_in_ABC]),
           param = all_sumstats[,1:2],
           sumstat = all_sumstats[,VAF_categories_to_include_in_ABC],
           tol = 0.05,
           transf = c("log","log"),
           method = "neuralnet")
  
  return(as.data.frame(res$adj.values)%>%mutate(exp_ID=Exp_ID$exp_ID))
})%>%dplyr::bind_rows()%>%
  left_join(ref_df%>%dplyr::select(Sample,Age),by=c("exp_ID"="Sample"))%>%
  dplyr::filter(!exp_ID%in%c("8pcw","18pcw"))
```

Perform the LMER on this "no non-synonymous mutation" results.

```{r}
lme.gens_by_age.syn<-lme4::lmer(total_generations~Age+(1|exp_ID),data=abc_syn_res_df)
summary(lme.gens_by_age.syn)
mtDNA_CN*365/lme.gens_by_age.syn@beta[2]
mtDNA_CN*365/confint(lme.gens_by_age.syn)["Age",]
```

Plot these results.

```{r fig.width=5,fig.height=4}
abc_synres_plot_total_generations<-abc_syn_res_df%>%
  dplyr::filter(!exp_ID%in%c("8pcw","18pcw"))%>%
  mutate(exp_ID=factor(exp_ID,levels=ref_df$Sample[order(ref_df$Age)]))%>%
  ggplot(aes(x=Age,y=total_generations))+
  geom_point(aes(col=exp_ID),alpha=0.05,size=0.25)+
  scale_y_continuous(limits=c(0,1700))+
  geom_abline(slope=lme.gens_by_age.syn@beta[2],intercept = lme.gens_by_age.syn@beta[1],linetype=1)+
  scale_color_manual(values=Individual_cols[-c(1:2)])+
  theme_classic()+
  guides(colour=guide_legend(override.aes = list(alpha=1)))+
  labs(x="Age",y="Total WF generations\n(Posterior distribution from ABC)",col="")+
  my_theme+
  theme(legend.box.spacing = unit(0,"mm"),
        legend.key.size = unit(0.5,"mm"),
        legend.spacing = unit(0,"mm"),
        legend.box.margin = margin(c(0,0,0,0)),
        legend.text = element_text(margin = margin(t=0)))

abc_synres_plot_total_generations+geom_point(aes(col=exp_ID),alpha=0.1,size=0.5)+my_rmd_theme+guides(colour=guide_legend(override.aes = list(alpha=1,size=1)))+
  theme(legend.box.spacing = unit(0,"mm"),
        legend.key.size = unit(1,"mm"),
        legend.spacing = unit(0.5,"mm"),
        legend.box.margin = margin(c(0,0,0,0)),
        legend.text = element_text(margin = margin(t=0)))

save_plot(filename=paste0(figures_dir,"abc_synres_plot_total_generations.pdf"),abc_synres_plot_total_generations,width=2,height=2)
```

### Compare the drift rate estimates between the full mutation set and the synonymous/ non-coding set

This shows that the estimates are very similar with highly overlapping confidence intervals.
Interestingly, despite being a smaller dataset, the non-coding mutation set has narrower confidence intervals and slightly lower drift parameter estimates - though clearly this is not statistically significant.

```{r fig.width=2.5,fig.height=3}
lme.syn.CIs<-mtDNA_CN*365/confint(lme.gens_by_age.syn)["Age",]
lme.full.CIs<-mtDNA_CN*365/confint(lme.gens_by_age)["Age",]

lme_res_df<-data.frame(mutation_set=c("full","coding\nexcluded"),
           drift_parameter=c(mtDNA_CN*365/lme.gens_by_age@beta[2],mtDNA_CN*365/lme.gens_by_age.syn@beta[2]),
           uci=c(lme.full.CIs[1],lme.syn.CIs[1]),
           lci=c(lme.full.CIs[2],lme.syn.CIs[2]))

lme_res_df%>%
  ggplot(aes(x=mutation_set,y=drift_parameter,ymin=lci,ymax=uci))+
  geom_point()+
  geom_linerange()+
  theme_classic()+
  lims(y=c(0,20000))+
  my_rmd_theme+
  labs(x="Mutation set",y="Drift parameter")

```

