---
title: "mtDNA mutation phasing"
author: "Mike Spencer Chapman"
date: "`r Sys.Date()`"
output: html_document
---

## R Markdown

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = F)

#Set these file paths before running the script
#genomeFile is set in config.R
source(here::here("config.R")) #sets root_dir, genomeFile, project paths and my_markdown_theme
set_markdown_figure_defaults() #larger default figure size for on-screen reading
source(paste0(root_dir,"/data/mito_mutations_blood_functions.R"))

#Set the key file paths using the root dir
tree_file_paths = list.files(paste0(root_dir,"/data/tree_files"),pattern=".tree",full.names = T)
ref_file=paste0(root_dir,"/data/Samples_metadata_ref.csv")
nonblood_ref_file=paste0(root_dir,"/data/metadata/non_blood_metadata.xlsx")
colony_info_file=paste0(root_dir,"/data/metadata/colonyinfo_AX001_KX001_KX002_KX003_TX001_TX002_CB001.txt")
figures_dir=notebook_plots_dir("mtDNA_mut_phasing") #see save_plots in config.R
rebuttal_figs_dir=paste0(root_dir,"/plots/additional_plots/") #plots that do not appear in the manuscript figures
plots_dir=notebook_plots_dir("mtDNA_mut_phasing")

# read in the trinucleotide context data
mtdna_trinuc_freq_path=ifelse(Sys.info()['sysname']=="Darwin",paste0(root_dir,"/data/mtDNA_trinuc_freqs_coding_dloop_heavy_light.Rds"),"/lustre/scratch126/casm/team154pc/ms56/mito_mutations_blood/data/mtDNA_trinuc_freqs_coding_dloop_heavy_light.Rds")
mtdna_trinuc_freq <- readRDS(mtdna_trinuc_freq_path)

#Set the path of your local genome file (this is hg19)
#genomeFile is set in config.R

# set regions for coding and dloop regions - important for the mutation profiles
coding_region <- 577:16023
d_loop_region <- c(1:576,16024:16569)

#Set the basic plotting theme for ggplot2
my_markdown_theme<-theme(text = element_text(family="Helvetica"),
                axis.text = element_text(size = 5),
                axis.title = element_text(size=7),
                legend.text = element_text(size=5),
                legend.title = element_text(size=7),
                strip.text = element_text(size=7),
                legend.spacing = unit(1,"mm"),
                legend.key.size= unit(5,"mm"))+
  theme(legend.key.height=unit(3,"mm"),
        legend.title = element_text(size=8))


#Read in the mitochondrial copy number data
mito_cn=read.csv(paste0(root_dir,"/data/whole_genome_coverage_pileup_and_bedtools_annotated.csv"),header=T)%>%
  dplyr::rename("pileup_median_mtDNA_coverage"=pilup_median_mtDNA_coverage) #This column has a typo

#Import individual level metadata for the adult and foetal blood samples
ref_df<-readxl::read_excel(nonblood_ref_file) #Import metadata relating to all individuals studied
colony_info<-read.delim(colony_info_file) #Import colony-level data for the lymphoid dataset as there are multiple different cell types

#Define the dataframe for converting IDs between my IDs, and Andrew's for the different datasets
translate=data.frame(dataset=c("KY","HL","SO","PR","LM","NW","lymph","EM"),
                     al_ref=c("lung_organoid","colon","colon_ibd","muty_mutant","endometrium","blood_MPN","immune","blood_emily"))

muty_samples=c("PD44887","PD44888","PD44889","PD44890","PD44891")

#Define a set of 'black-listed' mutations - these are sets of artefacts that recurrently slip through the filters
#This is often due to haplotype-specific artefacts that map as SNVs
exclude_muts=c("MT_302_A_C","MT_311_C_T","MT_456_C_T","MT_567_A_C","MT_574_A_C","MT_8270_C_T","MT_16170_A_C","MT_16181_A_C","MT_16182_A_C","MT_16183_A_C","MT_16189_T_C")

#Set up objects for colour schemes and renaming tissues for plots
mut_type_cols<-c("#00468B","#925E9F","#925E9F","#810F7C","black","#E10C04","black")
names(mut_type_cols)<-c("Missense","Nonsense","Truncating","Stop_loss","Non-Coding","Synonymous","Overall")

name_conversion_vec=c("Normal blood","Lymphoid","MPN","Colon","IBD-affected\nColon","MUTYH-mutant\nColon","Endometrium","Bronchial\nepithelium")
names(name_conversion_vec)=c("blood","lymph","NW","HL","SO","PR","LM","KY")
tissue_cols<-c("#96e97c","#fd81c8","#145a6a","#65e6f9","#781486","#5f70cc","#aea2eb","#1d6d1f")
names(tissue_cols)<-name_conversion_vec
```



```{r import_data}

all_cohorts_plus_CML<-c("KY","HL","SO","PR","LM","NW","CML","lymph","blood")
all_cohorts<-c("KY","HL","SO","PR","LM","NW","lymph","blood")

all_mito_datasets<-lapply(all_cohorts_plus_CML,function(dataset) {
  
  dataset_mito_data<-readRDS(paste0(root_dir,"/data/nonblood/mito_mutation_data_",dataset,".RDS"))
  CN_correlating_muts<-readRDS(paste0(root_dir,"/data/nonblood/CN_correlation_",dataset,".RDS"))
  
  rho_cut_off=0
  
  #Deal with the germline mutations (if not done already)
  dataset_mito_data<-Map(list=dataset_mito_data,exp_ID=names(dataset_mito_data),function(list,exp_ID) {
    if(is.null(list$germline_muts)) {
      list$germline_muts<-identify_germline(matrices=list$matrices,threshold=0.9)
      cat(list$germline_muts,sep="\n")
      list$matrices<-reverse_germline(matrices=list$matrices,threshold=0.9)
    }
    return(list)
  })
  
  #Sort out the names of the mutCN column
  dataset_mito_data<-lapply(dataset_mito_data,function(list) {
    colnames(list$matrices$implied_mutCN)<-stringr::str_split(colnames(list$matrices$implied_mutCN),pattern = '\\.\\.\\.',simplify=T)[,1]
    return(list)
  })
  
  #Add the CN correlating muts vector to each patient (slightly inefficient but programatically more straight-forward)
  dataset_mito_data<-lapply(dataset_mito_data,function(list) {
    list$CN_correlating_muts<-CN_correlating_muts
    return(list)
  })
  
  
  #cat("Adding the filtered VAF matrix.",sep="\n")
  
  #Annotate specific mutations with their most likely signature using the 'sig_ref' dataframe
  dataset_mito_data<-Map(list=dataset_mito_data,this_exp_ID=names(dataset_mito_data),function(list,this_exp_ID) {
    
    #cat(this_exp_ID,sep="\n")
    
    if(dataset=="lymph") {list$tree$tip.label<-unique(list$sample_shearwater_calls$sampleID)}
    
    mutCN_cutoff=25 #If the mitochondrial copy number is over 25, retain mutation even if is in the "CN correlating muts" list
    CN_correlating_mut_removal_mat=list$matrices$implied_mutCN>mutCN_cutoff|(matrix((!rownames(list$matrices$vaf)%in%CN_correlating_muts),ncol=1)%*%matrix(rep(1,ncol(list$matrices$vaf)),nrow=1))
    vaf.filt<-(list$matrices$vaf[,list$tree$tip.label]*list$matrices$SW[,list$tree$tip.label][,list$tree$tip.label]*CN_correlating_mut_removal_mat[,list$tree$tip.label])[!grepl("DEL|INS",rownames(list$matrices$vaf))&
                                                                                                                                                                            list$rho_vals>rho_cut_off&
                                                                                                                                                                            !rownames(list$matrices$vaf)%in%exclude_muts,]
    
    list$matrices$CN_correlating_mut_removal_mat<-CN_correlating_mut_removal_mat
    list$matrices$vaf.filt<-vaf.filt
    return(list)
  })
  
  if(dataset=="lymph") {
    dataset_mito_data<-Map(list=dataset_mito_data,exp_ID=names(dataset_mito_data),function(list,exp_ID) {
      if(!exp_ID=="TX001") {list$het_oocyte_muts<-c()}
      return(list)
    })
  }
  return(dataset_mito_data)
})
names(all_mito_datasets)<-all_cohorts_plus_CML
```

## DEFINE BESPOKE FUNCTIONS FOR THIS SCRIPT

The key principle is included in the "check_mtDNA_mut_phasing" function.
This looks at mutation pairs and does two tests between them:

Test 1: Do the vafs of the two mutations sum to >1 in any individual cell? If so, this is strong evidence that they must be on the SAME mtDNA molecule (the same haplotype)
Test 2: is there evidence of one mutation being 'sub-clonal' to the other i.e. one VAF is always at or equal to the other VAF. If there are some examples of mut1 VAF being > mut2 VAF, and other examples of mut2 VAF > mut1 VAF, this suggests that their VAFs are independent i.e. they are on DIFFERENT mtDNA molecules (different haplotypes).

Many mutation pairs will not meet either of these tests, and therefore their relationship to each other remains inconclusive from our data.

```{r define_custom_functions}
# The key function to check the phasing between two mutations
check_mtDNA_mut_phasing=function(mut1,mut2,vaf_mat,tol1=0.03,tol2=0.03) {
  
  if(!mut1%in%rownames(vaf_mat)|!mut2%in%rownames(vaf_mat)) {stop(cat("Either/ both mutations not found in the vaf matrix. Please check mutation references."))}
  if(mut1==mut2) {stop(return(NA))}
    
  #Perform test 1: Do the vafs of the two mutations sum to >1 in any individual cell
  sum_of_vaf<-sum(colSums(vaf_mat[c(mut1,mut2),])>(1+tol1))
  test1_evidence=max(colSums(vaf_mat[c(mut1,mut2),])-1) #This is the maximum value by which the sum of VAFs exceeds 1
  same_haplo_res<-sum_of_vaf>0
  
  #Perform test 2: is there evidence of one mutation being 'sub-clonal' to the other i.e. one VAF is always at or equal to the other VAF
  mut1_higher=sum((vaf_mat[mut1,]-vaf_mat[mut2,])>tol2)
  mut2_higher=sum((vaf_mat[mut2,]-vaf_mat[mut1,])>tol2)
  test2_evidence<-min(max((vaf_mat[mut1,]-vaf_mat[mut2,])),max((vaf_mat[mut2,]-vaf_mat[mut1,]))) #This is the minimum value by which, for each mutation, the maximum by which the VAF of that mutation is exceeded by that of the other in a given cell.
  diff_haplo_res<-mut1_higher>0 & mut2_higher>0
  
  if(same_haplo_res&diff_haplo_res) {
    if(test1_evidence>test2_evidence) {
      diff_haplo_res<-F
    } else {
      same_haplo_res<-F
    }
  }
  
  res<-ifelse(same_haplo_res&!diff_haplo_res,"Same haplotype",ifelse(diff_haplo_res&!same_haplo_res,"Different haplotype",ifelse(same_haplo_res & diff_haplo_res,"Inconsistent","Inconclusive")))
  
  if(res=="Same haplotype") {
    mut1_mean_vaf<-mean(unlist(vaf_mat[mut1,]))
    mut2_mean_vaf<-mean(unlist(vaf_mat[mut2,]))
    
    res<-ifelse(mut1_mean_vaf>mut2_mean_vaf,"Ancestral","Descendant")
  }
  return(res)
}

get_descendant_mut_names=function(tree,node) {
  descendant_node_numbers<-phangorn::Descendants(tree,node=node,type = "all")
  descendant_tips<-descendant_node_numbers[descendant_node_numbers<=length(tree$tip.label)]
  descendant_nodes<-descendant_node_numbers[descendant_node_numbers>length(tree$tip.label)]
  descendant_muts<-c(tree$node.label[descendant_nodes-length(tree$tip.label)],tree$tip.label[descendant_tips])
  return(descendant_muts)
}

get_ancestral_mut_names=function(tree,node) {
  ancestral_node_numbers<-phangorn::Ancestors(tree,node=node,type = "all")
  ancestral_muts<-tree$node.label[ancestral_node_numbers-length(tree$tip.label)]
  return(ancestral_muts)
}

get_direct_descendants<-function(phasing_mat,mut) {
  descendant_muts<-colnames(phasing_mat)[phasing_mat[mut,]=="Ancestral"]
  descendant_muts<-descendant_muts[!is.na(descendant_muts)]
  sub_phasing_mat<-phasing_mat[descendant_muts,descendant_muts,drop=F]
  rownames(sub_phasing_mat)[rowSums(sub_phasing_mat=="Descendant",na.rm = T)==0]
}

get_all_descendants<-function(phasing_mat,mut) {
  descendant_muts<-colnames(phasing_mat)[phasing_mat[mut,]=="Ancestral"]
  return(descendant_muts[!is.na(descendant_muts)])
}

rename_node=function(tree,tip_set,new_name) {
  matches_sub_tree=sapply(unique(tree$edge[,1]),function(node) {
    descendant_nodes<-phangorn::Descendants(tree,node=node,type="children")
    if(length(descendant_nodes)!=length(tip_set)) {stop(return(F))}
    all(tree$tip.label[descendant_nodes] == tip_set)
  })
  mut_node=unique(tree$edge[,1])[matches_sub_tree]
  tree$node.label[mut_node-length(tree$tip.label)]<-new_name
  return(tree)
}

#Build a tree based on these phasing results
#This is an Iterative function to build a genotype tree from the phasing results
build_mito_haplotype_tree=function(tree=NULL,muts,vaf_mat,phasing_mat) {
  
  if(is.null(tree)) {
    cat("Initiating tree",sep="\n")
    tree<-ape::stree(length(muts),type="star",tip.label=muts)
    tree$edge.length<-rep(1,length(muts))
    tree$node.label<-"ROOT"
  }
  
  for(mut in muts) {
    cat(mut,sep="\n")
    direct_descendant_muts<-get_direct_descendants(phasing_mat,mut)
    if(length(direct_descendant_muts)==0) {next}
    
    #Create subtree of the direct descendants of this mut
    if(length(direct_descendant_muts)==1) {
      #If there is only 1 descendant mutation need to create a dummy blank tip, otherwise it will get collapsed
      sub_tree<-stree(2,type="star",tip.label=c(direct_descendant_muts,""))
      sub_tree$edge.length<-c(1,0) #Add edge lengths
    } else {
      sub_tree<-stree(length(direct_descendant_muts),type="star",tip.label=direct_descendant_muts)
      sub_tree$edge.length<-rep(1,length(direct_descendant_muts)) #Add edge lengths
    }
    
    #Bind this to the old mutation
    new_tree<-bind.tree(x = tree,y=sub_tree,where=which(tree$tip.label==mut))
    
    #Label the newly created node with the mut_ref
    new_tree<-rename_node(new_tree,tip_set = sub_tree$tip.label,new_name = mut)
    
    #Remaining muts to include
    remaining_muts<-c(mut,get_all_descendants(phasing_mat = phasing_mat,mut = mut))
    sub_phasing_mat<-phasing_mat[remaining_muts,remaining_muts,drop=F]
    tree<-build_mito_haplotype_tree(tree=new_tree,muts=direct_descendant_muts,vaf_mat=vaf_mat,phasing_mat = sub_phasing_mat)
  }
  return(tree)
}
```

# Run the function over all the mutation data using the above functions

```{r run_mut_phasing_function,fig.height=22.5,fig.width=7.5}
temp=lapply(all_cohorts_plus_CML,function(cohort) {
  cat(cohort,sep="\n")
  dataset_mito_data<-all_mito_datasets[[cohort]]
  dataset_samples<-names(dataset_mito_data)
  temp2=lapply(dataset_samples,function(exp_ID) {
    cat(exp_ID,sep="\n")
    sample_mito_data<-dataset_mito_data[[exp_ID]]
    vaf.filt<-sample_mito_data$matrices$vaf.filt
    
    #Get mut_refs of mutations that are >3% in at least one sample
    high_vaf_mut_refs<-rownames(vaf.filt)[rowSums(vaf.filt>0.03)>0]
    
    #Set file paths for saving
    phasing_file_dir=paste0(root_dir,"/mut_phasing_output/")
    phasing_mat_path<-paste0(phasing_file_dir,"phasing_mat_",exp_ID,".Rds")
    tree_file_path<-paste0(phasing_file_dir,"tree_mtDNA_genotype_",exp_ID,".tree")
    tree_pdf_path<-paste0(phasing_file_dir,"genotype_tree_",exp_ID,".pdf")
    
    force_rerun=F
    if(file.exists(phasing_mat_path)&!force_rerun) {
      mat<-readRDS(phasing_mat_path)
    } else {
      mat=matrix(NA,nrow=length(high_vaf_mut_refs),ncol = length(high_vaf_mut_refs),dimnames = list(high_vaf_mut_refs,high_vaf_mut_refs))
      
      switcheroo=c("Ancestral","Descendant")
      names(switcheroo)<-rev(switcheroo)
      
      #Build the phasing matrix
      for(i in 1:nrow(mat)) {
        cat(i,sep="\t")
        for(j in 1:ncol(mat)) {
          if(i>=j) {
            next
          } else {
            mat[i,j]<-check_mtDNA_mut_phasing(mut1=rownames(mat)[i],mut2=rownames(mat)[j],vaf_mat=vaf.filt)
            mat[j,i]<-ifelse(mat[i,j]%in%c("Ancestral","Descendant"),switcheroo[mat[i,j]],mat[i,j])
          }
        }
      }
      
      #Save the phasing mat (this takes a long time to make)
      saveRDS(mat,file=phasing_mat_path)
    }
    
    #Start with all the mutations that aren't descendants of any other branches (i.e. that must therefore come from the 'root' genotype)
    muts_ancestral<-rownames(mat)[rowSums(mat=="Descendant",na.rm=T)==0]
    
    #Then build the tree from here & save
    tree<-build_mito_haplotype_tree(muts=muts_ancestral,vaf_mat=vaf.filt,phasing_mat=mat)
    write.tree(tree,file=tree_file_path)
    
    #plot the genotype phylogeny
    plot(tree,show.node.label=T,cex=0.3)
    
    #Now save it for later scrutiny
    save_pdf(tree_pdf_path,height=15,width=4)
    plot(tree,show.node.label=T,cex=0.3)
    dev.off()
    
  })
})

```

