---
title: "mtDNA Mutational Analysis - blood"
author: "Mike Spencer Chapman"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load packages (and install if they are not installed 
cran_packages=c("devtools","ape","stringr","dplyr","tidyr","ggplot2","gridExtra","phylosignal","ggpubr")
bioconductor_packages=c("MutationalPatterns","BSgenome","BSgenome.Hsapiens.UCSC.hg19","TxDb.Hsapiens.UCSC.hg19.knownGene")

for(package in cran_packages){
  if(!require(package, character.only=T,quietly = T, warn.conflicts = F)){
    install.packages(as.character(package),repos = "http://cran.us.r-project.org")
    library(package, character.only=T,quietly = T, warn.conflicts = F)
  }
}
if (!require("BiocManager", quietly = T, warn.conflicts = F))
  install.packages("BiocManager")
for(package in bioconductor_packages){
  if(!require(package, character.only=T,quietly = T, warn.conflicts = F)){
    BiocManager::install(as.character(package))
    library(package, character.only=T,quietly = T, warn.conflicts = F)
  }
}

if(!require("dndscv", character.only=T,quietly = T, warn.conflicts = F)){
  devtools::install_github("im3sanger/dndscv")
  library("dndscv",character.only=T,quietly = T, warn.conflicts = F)
}

```

Set file paths.
```{r}
options(stringsAsFactors = F)

#Set these file paths before running the script
genomeFile="~/Documents/Reference_files/hs37d5.fa"
root_dir="~/R_work/mito_mutations"
source(paste0(root_dir,"/data/mito_mutations_blood_functions.R"))

#Set the key file paths using the root dir
tree_file_paths = list.files(paste0(root_dir,"/data/tree_files"),pattern=".tree",full.names = T)
ref_file=paste0(root_dir,"/data/Samples_metadata_ref.csv")
figures_dir=paste0(root_dir,"/figures/")
rebuttal_figs_dir=paste0(root_dir,"/rebuttal_plots/")

#Set the basic plotting theme for ggplot2
my_theme<-theme(text = element_text(family="Helvetica"),
                axis.text = element_text(size = 5),
                axis.title = element_text(size=7),
                legend.text = element_text(size=5),
                legend.title = element_text(size=7),
                strip.text = element_text(size=7),
                legend.spacing = unit(1,"mm"),
                legend.key.size= unit(5,"mm"))+
  theme(legend.key.height=unit(3,"mm"),
        legend.title = element_text(size=8))

#Read in the mitochondrial copy number data
mito_cn=read.csv(paste0(root_dir,"/data/whole_genome_coverage_pileup_and_bedtools_annotated.csv"),header=T)

#Combine the sample level metadata info for the adult and foetal blood samples
sample_level_metadata_EM<-read.csv(paste0(root_dir,"/data/EM_sample_level_metadata.csv"))%>%dplyr::rename("exp_ID"="donor_id","Sample"="PDID","Cell_type"="cell_type","coverage"="mean_depth")
sample_level_metadata_foetal<-read.csv(paste0(root_dir,"/data/foetal_sample_level_metadata.csv"))%>%mutate(Sample=paste(Donor_ID,"hum",sep="_"))%>%dplyr::select(-Donor_ID,-Percentage)
sample_level_metadata<-dplyr::bind_rows(sample_level_metadata_EM,sample_level_metadata_foetal)

phenotype_data_EM<-read.csv(paste0(root_dir,"/data/Summary_pheno_pdid.csv"),stringsAsFactors = F)%>%
  dplyr::rename("exp_ID"=Donor_ID,"Sample"=PDID)

#Import individual level metadata for the adult and foetal blood samples
ref_df=read.csv(ref_file)%>%filter(Dataset!="Lymphocyte")
Individual_cols=RColorBrewer::brewer.pal(12,"Paired")
names(Individual_cols)<-ref_df$Sample[order(ref_df$Age)]

mito_cn<-mito_cn%>%
  left_join(sample_level_metadata,by="Sample")%>%
  left_join(ref_df%>%dplyr::rename("exp_ID"=Sample))
mito_cn$exp_ID<-factor(mito_cn$exp_ID,levels=ref_df$Sample[order(ref_df$Age)]) #Make the exp_ID a factor, with levels increasing by individual age
mito_cn<-left_join(mito_cn,phenotype_data_EM)

#Now import the mitochondrial mutation data
mito_data_file=paste0(root_dir,"/data/mito_data.Rds")
mito_data<-readRDS(mito_data_file)
CN_correlating_muts<-readRDS(paste0(root_dir,"/data/CN_correlation.RDS"))
exclude_muts=c("MT_302_A_C","MT_311_C_T","MT_567_A_C","MT_574_A_C","MT_16181_A_C","MT_16182_A_C","MT_16183_A_C","MT_16189_T_C") #These are the ones to exclude for the endometrial analysis

mut_type_cols<-c("#00468B","#925E9F","#925E9F","#810F7C","black","#E10C04","black")
names(mut_type_cols)<-c("Missense","Nonsense","Truncating","Stop_loss","Non-Coding","Synonymous","Overall")

```

## MITOCHONDRIAL COVERAGE STATISTICS

Assess the mtDNA coverage across samples.
This is a function of:
(1) the mtDNA copy number within the colony/ sample
(2) the overall WGS coverage of the sample

```{r}
#Generate the coverage histograms
coverage.plot<-mito_cn%>%
  dplyr::filter(Study%in%ref_df$Canapps.project & Sample%in% unlist(lapply(mito_data,function(list) list$tree$tip.label)))%>%
  ggplot(aes(x=bedtools_mtDNA_coverage))+
  geom_histogram(fill="lightblue",col="black",linewidth=0.2)+
  scale_x_log10(labels=scales::label_comma())+
  facet_wrap(~exp_ID,scales = "free_y")+
  theme_bw()+
  my_theme+
  theme(axis.text.x = element_text(angle=90),strip.text.x = element_text(size=6,margin = margin(1,0,1,0, "mm")))+
  labs(x="Mean mitochondrial DNA coverage",y="Count")
coverage.plot
ggsave(filename=paste0(figures_dir,"Supp_Figure_01/Coverage_plot.pdf"),coverage.plot,width=4,height=3)
```


```{r}
median.coverage.plot<-mito_cn%>%
  dplyr::filter(Study%in%ref_df$Canapps.project & Sample%in% unlist(lapply(mito_data,function(list) list$tree$tip.label)))%>%
  ggplot(aes(x=pilup_median_mtDNA_coverage))+
  geom_histogram(fill="lightblue",col="black",linewidth=0.2)+
  scale_x_log10(labels=scales::label_comma())+
  facet_wrap(~exp_ID,scales = "free_y",nrow=2)+
  theme_bw()+
  my_theme+
  theme(axis.text.x = element_text(angle=90),strip.text.x = element_text(size=6,margin = margin(1,0,1,0, "mm")))+
  labs(x="Median mitochondrial DNA coverage",y="Count")
median.coverage.plot
ggsave(filename=paste0(figures_dir,"Supp_Figure_01/Coverage_median_plot.pdf"),median.coverage.plot,width=7,height=2)
```

A ridges plot is perhaps better here to compare.
```{r coverage_ridges,fig.height=3,fig.width=4}
median.coverage.ridges.plot<-mito_cn%>%
  dplyr::filter(Study%in%ref_df$Canapps.project & Sample%in% unlist(lapply(mito_data,function(list) list$tree$tip.label)))%>%
  ggplot(aes(x=pilup_median_mtDNA_coverage,y=exp_ID))+
  ggridges::geom_density_ridges(linewidth=0.2,fill="lightblue")+
  scale_x_log10(labels=scales::label_comma())+
  theme_bw()+
  my_theme+
  theme(axis.text.x = element_text(angle=90),strip.text.x = element_text(size=6,margin = margin(1,0,1,0, "mm")))+
  labs(x="Median mitochondrial DNA coverage",y="")
median.coverage.ridges.plot
ggsave(filename=paste0(figures_dir,"Supp_Figure_01/median.coverage.ridges.plot.pdf"),median.coverage.ridges.plot,width=3.3,height=2.5)
```

Uniformity across the mtDNA genome is also a key marker of coverage quality.
We use a very stringent marker of uniformity - the proportion of genome with >80% of the mean.

```{r uniformity_ridges,fig.height=3,fig.width=4}
uniformity_ridges.plot<-mito_cn%>%
  dplyr::filter(Study%in%ref_df$Canapps.project & Sample%in% unlist(lapply(mito_data,function(list) list$tree$tip.label)))%>%
  ggplot(aes(x=perc_over0.8,y=exp_ID))+
  ggridges::geom_density_ridges(linewidth=0.2,fill="lightblue")+
  scale_x_log10()+
  theme_bw()+
  my_theme+
  theme(axis.text.x = element_text(angle=90),strip.text.x = element_text(size=6,margin = margin(1,0,1,0, "mm")))+
  labs(x="Coverage uniformity\n(proportion of mtDNA genome with coverage >80% of mean)",y="")
uniformity_ridges.plot
ggsave(filename=paste0(figures_dir,"Supp_Figure_01/Uniformity_perc80_ridges_plot.pdf"),uniformity_ridges.plot,width=3.3,height=2.5)
```

Print coverage summary statistics
```{r}
mito_cn%>%
  dplyr::filter(Study%in%ref_df$Canapps.project & Sample%in% unlist(lapply(mito_data,function(list) list$tree$tip.label)))%>%
  group_by(exp_ID)%>%
  summarise(samples=n(),mean_mtDNA_coverage=mean(bedtools_mtDNA_coverage),Under_1000X=sum(bedtools_mtDNA_coverage<1000))

mito_cn%>%
  dplyr::filter(Study%in%ref_df$Canapps.project & Sample%in% unlist(lapply(mito_data,function(list) list$tree$tip.label)))%>%
  summarise(mean_mtDNA_coverage=mean(bedtools_mtDNA_coverage))

```

## Mitochondrial DNA copy number
Get some basic MITOCHONDRIAL COPY NUMBER STATISTICS
```{r}
#Vector to convert the foetal cell type codes into basic 'HSC'/ 'Progenitor' types
convert_vec=c("HSC","HSC","HPC","HPC","HPC","HPC")
names(convert_vec)=c("H","HSC","C","M","HSPC","Progenitor")

mtDNA_by_pheno_and_ID_summary<-mito_cn%>%
  filter(!is.na(Phenotype) & Cell_type=="HSC")%>%
  group_by(exp_ID,Phenotype)%>%
  summarise(n=n(),
            mean_mtDNA_genomes=mean(bedtools_mtDNA_genomes),
            median_mtDNA_genomes=median(bedtools_mtDNA_genomes))

mtDNA_by_pheno_summary<-mito_cn%>%
  filter(!is.na(Phenotype) & Cell_type=="HSC")%>%
  group_by(Phenotype)%>%
  summarise(n=n(),
            mean_mtDNA_genomes=mean(bedtools_mtDNA_genomes),
            median_mtDNA_genomes=median(bedtools_mtDNA_genomes))
```

Review the average copy number of colonies with different phenotypes relative to the average copy number of erythroid colonies

```{r fig.height=2.5,fig.width=3.5}
pheno_levels=c("Ery","EryMy","Gran","MyGran","MyMono","Mono","NKMy")
mtDNA_by_pheno_and_ID_summary%>%
  dplyr::select(-n)%>%
  pivot_wider(id_cols=c("exp_ID"),names_from="Phenotype",values_from="median_mtDNA_genomes")%>%
  mutate(across(all_of(pheno_levels),.fns = function(x) x/Ery))


mito_cn_by_pheno<-mito_cn%>%
  filter(!is.na(Phenotype) & Cell_type=="HSC")%>%
  dplyr::select(exp_ID,Sample,bedtools_mtDNA_genomes,Phenotype)%>%
  ggplot(aes(x=factor(Phenotype,levels=pheno_levels),y=bedtools_mtDNA_genomes))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(aes(col=factor(exp_ID,levels=ref_df$Sample[order(ref_df$Age)])),width=0.2,height=0,alpha=0.3,size=0.025)+
  scale_y_log10(limits=c(50,3000))+
  theme_bw()+
  scale_color_manual(values = Individual_cols[-c(1,2)])+
  guides(col=guide_legend(override.aes = list(alpha=1,size=0.5)))+
  labs(x="Colony cell composition",y="mtDNA copies per cell",col="Individual")+
  geom_text(data=mtDNA_by_pheno_summary,aes(label=paste0("µ = ",round(mean_mtDNA_genomes)),x=Phenotype,y=75),size=1.6,inherit.aes = F)+
  my_theme+
  theme(#axis.text.x = element_text(angle=90),
        strip.text.x = element_text(size=7,margin = margin(t=1,r=0,b=1,l=0,unit="mm")),
        legend.key.size = unit(0,"mm"),
        legend.position="right")
mito_cn_by_pheno
ggsave(filename=paste0(figures_dir,"Additional_plots/mito_cn_by_pheno.pdf"),mito_cn_by_pheno,width=4,height=2)
```


```{r fig.height=2,fig.width=4}
mito_cn_by_pheno_all_HSCprog<-mito_cn%>%
  filter(!is.na(Phenotype))%>%
  mutate(Cell_type=ifelse(Cell_type=="Progenitor","HPC",Cell_type))%>%
  dplyr::select(exp_ID,Cell_type,Sample,bedtools_mtDNA_genomes,Phenotype)%>%
  ggplot(aes(x=factor(Cell_type),y=bedtools_mtDNA_genomes,col=factor(Cell_type)))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width=0.3,height=0,alpha=0.1,size=0.1)+
  ggpubr::stat_compare_means(aes(label = paste0("p=", ..p.format..)), label.x=1.0,label.y=3.7,method="wilcox.test",size=1.8,fontface="italic")+
  scale_color_brewer(palette = "Set2")+
  scale_y_log10(limits=c(8,20000),breaks=c(10,100,1000,10000))+
  facet_grid(~factor(Phenotype,levels=pheno_levels))+
  theme_bw()+
  labs(x="",col="Founder cell\nphenotype",y="mtDNA copies per cell")+
  my_theme+
  theme(axis.text.x = element_text(angle=90),strip.text.x = element_text(size=6,margin = margin(t=1,r=0,b=1,l=0,unit="mm")),legend.position="none",legend.title = element_text(size=7))
mito_cn_by_pheno_all_HSCprog
ggsave(filename=paste0(figures_dir,"Figure_01/mito_cn_by_pheno_HSCprog.pdf"),mito_cn_by_pheno_all_HSCprog,width=4,height=2)

#Generate the table of average CN by cell type - these values are overlaid onto the above plot
mito_cn%>%
  filter(!is.na(Phenotype))%>%
  mutate(Cell_type=ifelse(Cell_type=="Progenitor","HPC",Cell_type))%>%
  dplyr::select(exp_ID,Cell_type,Sample,bedtools_mtDNA_genomes,Phenotype)%>%
  group_by(Cell_type,Phenotype)%>%
  summarise(median_CN=median(bedtools_mtDNA_genomes))

```

Generate a plot of mtDNA copy number across samples, grouped by individual.

```{r fig.height=2,fig.width=7}
Individual_means_df<-mito_cn%>%
  mutate(Cell_type=convert_vec[Cell_type])%>%
  filter(Sample%in%unlist(lapply(mito_data,function(list) list$tree$tip.label))&
           !is.na(bedtools_mtDNA_genomes)&
           #Cell_type=="HSC"&
           !is.na(exp_ID))%>%
  group_by(exp_ID)%>% 
  summarise(mean=mean(bedtools_mtDNA_genomes,na.rm=T),median=median(bedtools_mtDNA_genomes,na.rm=T),sd=sd(bedtools_mtDNA_genomes,na.rm=T),IQR_lower=quantile(bedtools_mtDNA_genomes,0.25),IQR_upper=quantile(bedtools_mtDNA_genomes,0.75))
Individual_means_df$exp_ID<-factor(Individual_means_df$exp_ID,levels=ref_df$Sample[order(ref_df$Age)])

studies_to_include=ref_df$Canapps.project
mtDNA.copy.number.logscale<-mito_cn%>%
  mutate(Cell_type=convert_vec[Cell_type])%>%
  filter(Sample%in%unlist(lapply(mito_data,function(list) list$tree$tip.label))&
           !is.na(bedtools_mtDNA_genomes)&
           !is.na(exp_ID))%>%
  ggplot(aes(x=factor(Sample,levels=mito_cn%>%dplyr::filter(Study%in%studies_to_include)%>%arrange(bedtools_mtDNA_genomes)%>%pull(Sample)%>%unique()),
             y=bedtools_mtDNA_genomes))+
  geom_point(size = 0.7, stroke = 0, shape = 16,col="black")+
  scale_y_log10(limits=c(8,20000),breaks=c(10,100,1000,10000))+
  theme_classic()+
  facet_grid(cols=vars(factor(exp_ID,levels = ref_df$Sample[order(ref_df$Age)])),scales="free",space = "free")+
  geom_hline(aes(yintercept=median),linetype=2,data=Individual_means_df,col="red",linewidth=0.5)+
  geom_text(aes(x=100,y=mean,label=paste0("tilde(x) == ",round(median))),nudge_y=+0.1,size=2,data=Individual_means_df,parse=T,inherit.aes = F)+
  labs(x="Individual",y="Mitochondrial copy number",col="Colony\nphenotype")+
  my_theme+
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.title = element_text(size=8),
        legend.key.height = unit(2,"mm"),
        strip.text.x = element_text(size=6),
        panel.spacing.x=unit(1, "mm"))+
  guides(colour = guide_legend(override.aes = list(size=2)))

mtDNA.copy.number.logscale

ggsave(filename = paste0(figures_dir,"Figure_01/mtDNA.copy.number.logscale.pdf"),plot=mtDNA.copy.number.logscale,height=2,width=7)
```

Generate the same plot, but now colour points by the colony phenotype. This is slightly difficult to interpret due to the multitude of colours!

```{r fig.height=2,fig.width=7}
phenotype_cols=c("red",RColorBrewer::brewer.pal(n=length(pheno_levels)-1,name = "Set2"),"darkgray") #Make the erythroid red, and 'unknown' dark grey
names(phenotype_cols)<-c(pheno_levels,"Unknown")
mtDNA.copy.number.logscale.by.phenotype<-mito_cn%>%
  mutate(Cell_type=convert_vec[Cell_type])%>%
  filter(Sample%in%unlist(lapply(mito_data,function(list) list$tree$tip.label))&
           !is.na(bedtools_mtDNA_genomes)&
           !is.na(exp_ID))%>%
  replace_na(replace=list(Phenotype="Unknown"))%>%
  ggplot(aes(x=factor(Sample,levels=mito_cn%>%dplyr::filter(Study%in%studies_to_include)%>%arrange(bedtools_mtDNA_genomes)%>%pull(Sample)%>%unique()),
             y=bedtools_mtDNA_genomes,
             col=Phenotype))+
  geom_point(size = 0.3, alpha=0.7,stroke = 0, shape = 16)+
  scale_y_log10(limits=c(8,20000),breaks=c(10,100,1000,10000))+
  scale_color_manual(values=phenotype_cols)+
  theme_classic()+
  facet_grid(cols=vars(factor(exp_ID,levels = ref_df$Sample[order(ref_df$Age)])),scales="free",space = "free")+
  geom_hline(aes(yintercept=median),linetype=2,data=Individual_means_df,col="red",linewidth=0.5)+
  geom_text(aes(x=100,y=mean,label=paste0("tilde(x) == ",round(median))),nudge_y=+0.1,size=2,data=Individual_means_df,parse=T,inherit.aes = F)+
  labs(x="Individual",y="Mitochondrial copy number",col="Colony\nphenotype")+
  my_theme+
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.title = element_text(size=8),
        legend.key.height = unit(2,"mm"),
        strip.text.x = element_text(size=6),
        panel.spacing.x=unit(1, "mm"))+
  guides(colour = guide_legend(override.aes = list(size=2)))

mtDNA.copy.number.logscale.by.phenotype

ggsave(filename = paste0(figures_dir,"Figure_01/mtDNA.copy.number.logscale.by.phenotype.pdf"),plot=mtDNA.copy.number.logscale.by.phenotype,height=2,width=7)
```


```{r fig.height=2,fig.width=3}
#Look in more details at the 18 pcw copy number by cell type & location
convert_vec2=c("HSC","MEP","CMP"); names(convert_vec2)=c("H","M","C")
convert_vec3=c("fBM","fLiver");names(convert_vec3)<-c("F","L")
foetal_18pcw_by_cell_type_loc<-mito_cn%>%
  filter(exp_ID=="18pcw")%>%
  mutate(Cell_type2=convert_vec[Cell_type])%>%
  mutate(Cell_type=convert_vec2[Cell_type],Tissue=convert_vec3[gsub("1|2","",Tissue.y)])%>%
  mutate(Cell_type_loc=paste(Tissue,Cell_type,sep="\n"))%>%
  ggplot(aes(x=factor(Cell_type_loc,levels=apply(expand.grid(convert_vec3,convert_vec2),1,paste,collapse="\n")),
             y=bedtools_mtDNA_genomes,
             col=Cell_type2))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(width=0.15,height=0,size=0.3,alpha=0.5)+
  scale_color_brewer(palette="Set2")+
  theme_bw()+
  my_theme+
  scale_y_log10(limits=c(8,20000),breaks=c(10,100,1000,10000))+
  theme(legend.title = element_text(size=7),legend.text = element_text(size=6))+
  labs(x="Cell type & location",y="mtDNA copies per cell",col="Founder cell\nphenotype")

foetal_18pcw_by_cell_type_loc
ggsave(filename=paste0(figures_dir,"Figure_01/foetal_18pcw_by_cell_type_loc.pdf"),foetal_18pcw_by_cell_type_loc,width=3,height=2)
```

### Assess for a relationship between age and mtDNA copy number

Plot the median mitochondrial CN by age

```{r fig.height=2,fig.width=3}
median_mtDNA_CN_by_age<-mito_cn%>%
  filter(Sample%in%unlist(lapply(mito_data,function(list) list$tree$tip.label))&
           !is.na(bedtools_mtDNA_genomes)&
           !is.na(exp_ID)&
           Cell_type=="HSC")%>%
  group_by(exp_ID)%>% 
  summarise(mean=mean(bedtools_mtDNA_genomes,na.rm=T),median=median(bedtools_mtDNA_genomes,na.rm=T),sd=sd(bedtools_mtDNA_genomes,na.rm=T))%>%
  left_join(ref_df,by=c("exp_ID"="Sample"))%>%
  ggplot(aes(x=Age,y=median,col=factor(exp_ID,levels=ref_df$Sample[order(ref_df$Age)])))+
  geom_smooth(aes(x=Age,y=median),method="lm",inherit.aes = F,col="black",size=0.5)+
  geom_point(alpha=0.8)+
  scale_color_manual(values = Individual_cols)+
  theme_bw()+
  scale_y_continuous(limits=c(0,1000))+
  labs(y="Median mitochondrial genomes per cell",col="Individual")+
  my_theme+
  theme(legend.key.height = unit(3,"mm"),legend.title=element_text(size=7),legend.text=element_text(size=6))
median_mtDNA_CN_by_age
ggsave(filename = paste0(figures_dir,"Supp_Figure_01/median_mtDNA_CN_by_age.pdf"),plot = median_mtDNA_CN_by_age,width=3.5,height=2)
```

Now do a formal regression to confirm if there is any kind of significant relationship (although not evident visually).
```{r}
CN_by_age_summary<-mito_cn%>%
  filter(Sample%in%unlist(lapply(mito_data,function(list) list$tree$tip.label))&
           !is.na(bedtools_mtDNA_genomes)&
           !is.na(exp_ID)&
           Cell_type=="HSC")%>%
  group_by(exp_ID)%>% 
  summarise(mean=mean(bedtools_mtDNA_genomes,na.rm=T),median=median(bedtools_mtDNA_genomes,na.rm=T),sd=sd(bedtools_mtDNA_genomes,na.rm=T))%>%
  left_join(ref_df,by=c("exp_ID"="Sample"))
lm.CN_by_age<-lm(median~Age,data=CN_by_age_summary)
summary(lm.CN_by_age)
confint(lm.CN_by_age)
```

### Test mitochondrial copy number for phylogenetic signal

How heritable is mitochondrial copy number? One way to look at this is to look for "phylogenetic signal".
This looks at whether the variation in copy number is explained by their lineage relationships, as defined by the phylogeny.

```{r mtDNA_CN_phylogenetic_signal,fig.height=10,fig.width=6}
if(!file.exists(paste0(root_dir,"/data/mito_cn_phylo.Rds"))) {
  mito_cn_phylo<-lapply(mito_data,function(list) {
    tree.no.ancestral=drop.tip(list$tree.ultra,"Ancestral")
    mito_cn_tree<-sapply(tree.no.ancestral$tip.label,function(sampleID) mito_cn%>%filter(Sample==sampleID&Study%in%studies_to_include)%>%slice_head(n=1)%>%pull(bedtools_mtDNA_genomes))
    tree.4d<-phylobase::phylo4d(tree.no.ancestral,tip.data=mito_cn_tree[tree.no.ancestral$tip.label])
    barplot.phylo4d(tree.4d)
    res<-phyloSignal(tree.4d)
    mut.crlg <- phyloCorrelogram(tree.4d)
    return(list(mito_cn=mito_cn_tree,tree4d=tree.4d,phyloSignal_res=res,phyloCor_res=mut.crlg))
  })
  saveRDS(mito_cn_phylo,file=paste0(root_dir,"/data/mito_cn_phylo.Rds"))
} else {
  mito_cn_phylo<-readRDS(paste0(root_dir,"/data/mito_cn_phylo.Rds"))
}

#Plot the mitochondrial copy number onto the trees (given the broad variation - plot as the log value)
par(mfrow=c(6,2))
Map(list=mito_data,mito_cn_phy=mito_cn_phylo,exp_ID=names(mito_data),function(list,mito_cn_phy,exp_ID){
  tree=plot_tree(list$tree.ultra,bars = log(mito_cn_phy$mito_cn),title=exp_ID)
})
```

Plot a histogram of the p-values to show that, while there are a couple <0.5, overall there is not a convincing signal.

```{r mtDNA_CN_phylogenetic_signal,fig.height=2,fig.width=2.5}
#Histogram of p-values
sapply(mito_cn_phylo,function(dat) dat[[3]]$pvalue["dt","Cmean"])%>%
  as.data.frame()%>%
  tibble::rownames_to_column()%>%
  dplyr::rename("exp_ID"=1,"p-value"=2)%>%
  ggplot(aes(x=`p-value`))+
  geom_histogram(breaks=seq(0,1,0.04),fill="lightblue",col="black",linewidth=0.2)+
  scale_x_continuous(breaks=seq(0,1,0.1))+
  scale_y_continuous(breaks=seq(0,2,1))+
  theme_classic()+
  my_theme+
  labs(x="Unadjusted p-values")

```

# Mitochondrial DNA mutations

Generate a tidy dataframe of samples, mutations and VAFs (representing the heteroplasmy in the cell population).

```{r}

mutCN_cutoff=25 #If the mitochondrial copy number is over 25, retain mutation even if is in the "CN correlating muts" list
df_tidy<-dplyr::bind_rows(Map(list=mito_data,exp_ID=names(mito_data),function(list,exp_ID) {
  if(is.null(list)){stop(return(NULL))}
  CN_correlating_mut_removal_mat=list$matrices$implied_mutCN>mutCN_cutoff|(matrix((!rownames(list$matrices$vaf)%in%CN_correlating_muts),ncol=1)%*%matrix(rep(1,ncol(list$matrices$vaf)),nrow=1))
  implied_mut_CN_tidy<-list$matrices$implied_mutCN%>%
    as.data.frame()%>%
    tibble::rownames_to_column(var="mut_ref")%>%
    dplyr::select(-global)%>%
    tidyr::gather(key="Sample",value="implied_mut_CN",-mut_ref)
  
  df_tidy<-(list$matrices$vaf*list$matrices$SW*(list$matrices$ML_Sig=="N1")*CN_correlating_mut_removal_mat)%>%
    as.data.frame()%>%
    tibble::rownames_to_column(var="mut_ref")%>%
    mutate(rho_val=list$rho_vals)%>%
    dplyr::select(-global)%>%
    tidyr::gather(key="Sample",value="vaf",-mut_ref,-rho_val)
  
  comb_tidy<-left_join(df_tidy,implied_mut_CN_tidy,by=c("Sample","mut_ref"))%>%
    dplyr::filter(!grepl("DEL|INS",mut_ref) & !mut_ref%in%exclude_muts)%>% #Exclude indels, and the 'black listed' mutations
    dplyr::filter(vaf>0)%>%
    mutate(exp_ID=exp_ID)
  return(comb_tidy)
  }))

#Filter the CN-correlating mutations - due to mis-mapping of nuclear reads.
#There may be some genuine mutations at these sites, in which case the implied mutation copy number will be much higher than ~2 (here an arbitrary threshold of 8 is applied)
df_tidy<-df_tidy%>%
  left_join(mito_cn%>%mutate(exp_ID=gsub("8pcw","8 pcw",exp_ID)))%>%
  mutate(implied_mut_CN=vaf*bedtools_mtDNA_genomes)%>%
  dplyr::filter(!(mut_ref%in%CN_correlating_muts & implied_mut_CN<=mutCN_cutoff))

head(df_tidy)

```

### Print headline statistics

Print basic stats of total mutation numbers called
```{r}
vaf_cut_off<-0
rho_cut_off<-0
sum(unlist(lapply(mito_data,function(list) {
  if(is.null(list)){stop(return(NULL))}
  vaf.filt<-(list$matrices$vaf*list$matrices$SW)[!grepl("DEL|INS",rownames(list$matrices$vaf))&
                                                   list$rho_vals>rho_cut_off&
                                                   !rownames(list$matrices$vaf)%in%exclude_muts,]
  total_pos=rowSums(vaf.filt>0,na.rm = T)
  return(sum(total_pos>0,na.rm=T))
})))
```

Total number of SNVs identified by shearwater - once CN correlating muts excluded
```{r}
sum(unlist(lapply(mito_data,function(list) {
  if(is.null(list)){stop(return(NULL))}
  CN_correlating_mut_removal_mat=list$matrices$implied_mutCN>mutCN_cutoff|(matrix((!rownames(list$matrices$vaf)%in%CN_correlating_muts),ncol=1)%*%matrix(rep(1,ncol(list$matrices$vaf)),nrow=1))
  vaf.filt<-(list$matrices$vaf*list$matrices$SW*CN_correlating_mut_removal_mat)[!grepl("DEL|INS",rownames(list$matrices$vaf))&
                                                                                  list$rho_vals>rho_cut_off&
                                                                                  !rownames(list$matrices$vaf)%in%exclude_muts,]
  total_pos=rowSums(vaf.filt>0,na.rm = T)
  return(sum(total_pos>0,na.rm=T))
})))
```

Total number of SNVs identified by shearwater - once CN correlating muts excluded and only N1 muts included
```{r}
sum(unlist(lapply(mito_data,function(list) {
  if(is.null(list)){stop(return(NULL))}
  CN_correlating_mut_removal_mat=list$matrices$implied_mutCN>mutCN_cutoff|(matrix((!rownames(list$matrices$vaf)%in%CN_correlating_muts),ncol=1)%*%matrix(rep(1,ncol(list$matrices$vaf)),nrow=1))
  vaf.filt<-(list$matrices$vaf*list$matrices$SW*(list$matrices$ML_Sig=="N1")*CN_correlating_mut_removal_mat)[!grepl("DEL|INS",rownames(list$matrices$vaf))&
                                                                                                               list$rho_vals>rho_cut_off&
                                                                                                               !rownames(list$matrices$vaf)%in%exclude_muts,]
  total_pos=rowSums(vaf.filt>0,na.rm = T)
  return(sum(total_pos>0,na.rm=T))
})))
```

Generate a dataframe of mean mutations VAFs of mutations. Though not particularly interesting.

```{r}
Individual_burden_means_df<-df_tidy%>%
  dplyr::filter(Sample!="global"&
                  #ML_Sig=="N1"&
                  !mut_ref%in%exclude_muts)%>%
  group_by(Sample)%>%
  dplyr::summarise(exp_ID=unique(exp_ID),n_mut=sum(vaf>vaf_cut_off))%>%
  tidyr::complete(Sample,fill=list(n_mut=0))%>%
  mutate(exp_ID=sapply(Sample,function(SampleID) {mito_cn$exp_ID[mito_cn$Sample==SampleID][1]}))%>%
  mutate(exp_ID=gsub("8 pcw","8pcw",exp_ID))%>%
  group_by(exp_ID)%>%
  summarise(mean=mean(n_mut))

head(Individual_burden_means_df)
```

Generate Fig. 1
This shows what proportion of samples have at least one mutation above the threshold value on the x-axis.

```{r fig.height=2,fig.width=3}
#Cycle through the different thresholds (from 0.01 to 0.99) and see how many mutations are above that cut-off in each sample.
#Then group by the different individuals and look at the proportion of samples with at >0 mutations at that cutoff.

samples_with_mut_df<-dplyr::bind_rows(lapply(seq(0.01,0.99,0.01),function(cut_off) {
  cut_off_df<-df_tidy%>%
    dplyr::filter(Sample!="global"&
                    #ML_Sig=="N1"&
                    !mut_ref%in%exclude_muts)%>%
    group_by(Sample)%>%
    dplyr::summarise(n_mut=sum(vaf>cut_off))%>%
    tidyr::complete(Sample,fill=list(n_mut=0))%>%
    mutate(exp_ID=sapply(Sample,function(SampleID) {mito_cn$exp_ID[mito_cn$Sample==SampleID][1]}))%>%
    group_by(exp_ID)%>%
    dplyr::summarise(mean=mean(n_mut),n_with_mut=sum(n_mut>0),n_samp=n())%>%
    mutate(prop_with_mut=n_with_mut/n_samp,cut_off=cut_off)
  return(cut_off_df)
}))

prop.of.samples.with.mut<-samples_with_mut_df%>%
  dplyr::mutate(exp_ID=gsub("8 pcw","8pcw",exp_ID))%>%
  ggplot(aes(x=cut_off,y=prop_with_mut,col=factor(exp_ID,levels = ref_df$Sample[order(ref_df$Age)])))+
  #geom_point()+
  geom_line(alpha=0.6)+
  theme_bw()+
  scale_y_continuous(limits=c(0,1),breaks=seq(0,1,0.2))+
  scale_color_brewer(palette="Paired")+
  labs(x="Cut-off",
       y=str_wrap("Proportion of samples with at least 1 mutation with VAF > cut-off",width=40),
       col="Individual")+
  my_theme+
  theme(legend.key.height = unit(3,"mm"),legend.title = element_text(size=7))

prop.of.samples.with.mut

ggsave(filename=paste0(figures_dir,"Figure_01/Sample_proportions_with_mut.pdf"),prop.of.samples.with.mut,width=3,height=2)

```

### Assess burdens using the sum of VAF measure

A good measure of the overall mutation burden is the "sum of VAF".
For example, if this is 1, it means that - on average - each mitochondrial genome has one mutation.

```{r}

sum_of_vaf_df<-Map(list=mito_data,Exp_ID=names(mito_data),function(list,Exp_ID){
  #sum the vaf of mutations that:
  #(1) Pass shearwater,
  #(2) are most likely signature N1 (the real mutatation signature),
  #(3) are not indels, specific artefacts or 'copy number correlating mutations' (i.e. most likely artefacts from mismapping of nuclear DNA or low-level contamination)
  CN_correlating_mut_removal_mat=list$matrices$implied_mutCN>mutCN_cutoff|(matrix((!rownames(list$matrices$vaf)%in%CN_correlating_muts),ncol=1)%*%matrix(rep(1,ncol(list$matrices$vaf)),nrow=1))
  df<-as.data.frame(colSums((list$matrices$vaf*list$matrices$SW*(list$matrices$ML_Sig=="N1")*(CN_correlating_mut_removal_mat))[!grepl("DEL|INS",rownames(list$matrices$vaf))&
                                                                                                !rownames(list$matrices$vaf)%in%exclude_muts,],na.rm = T))%>%
    tibble::rownames_to_column(var="Sample")%>%
    mutate(exp_ID=Exp_ID)%>%
    dplyr::rename(sum_of_vaf=2)%>%
    dplyr::filter(Sample!="global")
  return(df)
})%>%
  dplyr::bind_rows()%>%
  left_join(mito_cn%>%dplyr::select(Sample,Cell_type,Phenotype),by="Sample")%>%
  dplyr::mutate(Cell_type=convert_vec[Cell_type])

sum_of_vaf_summary<-sum_of_vaf_df%>%
  mutate(exp_ID=gsub("8 pcw","8pcw",exp_ID))%>%
  group_by(exp_ID)%>%
  summarise(mean=mean(sum_of_vaf,na.rm=T),median=median(sum_of_vaf,na.rm = T))

head(sum_of_vaf_summary)
```


```{r}
sum_of_vaf_df%>%
  left_join(mito_cn,by=c("Sample","exp_ID"))%>%
  dplyr::mutate(exp_ID=gsub("8 pcw","8pcw",exp_ID))%>%
  ggplot(aes(x=factor(Cell_type,levels = c("HSC","Progenitor")),col=Cell_type,y=sum_of_vaf))+
  ggpubr::stat_compare_means(aes(label = paste0("p=", ..p.format..)), label.x=1.0,label.y=3.7,method="wilcox.test",size=1.8,fontface="italic")+
  geom_point(alpha=0.1)+
  facet_grid(~factor(exp_ID,levels=ref_df$Sample[order(ref_df$Age)]),drop=T)+
  geom_boxplot()+
  theme_bw()+
  my_theme
```


```{r}
mean_sum_of_vaf_by_age<-sum_of_vaf_summary%>%
  left_join(ref_df,by=c("exp_ID"="Sample"))%>%
  ggplot(aes(x=Age,y=mean,col=factor(exp_ID,levels = ref_df$Sample[order(ref_df$Age)])))+
  geom_point()+
  theme_bw()+
  scale_color_brewer(palette="Paired")+
  labs(col="Individual",y="Mean mutation burden (sum of VAF)")+
  geom_smooth(aes(x=Age,y=mean),col="black",linewidth=0.5,method="lm",inherit.aes = F)+
  my_theme+
  theme(legend.key.height = unit(3,"mm"),legend.title=element_text(size=7),legend.text=element_text(size=5))

mean_sum_of_vaf_by_age

ggsave(filename=paste0(figures_dir,"Figure_01/mean_sum_of_vaf_by_age.pdf"),plot=mean_sum_of_vaf_by_age,height=2,width=3)
```


```{r}
lm.mean_mutburden_by_age=lm(mean~Age,data=sum_of_vaf_summary%>%
                              left_join(ref_df,by=c("exp_ID"="Sample")))
summary(lm.mean_mutburden_by_age)$coefficients
summary(lm.mean_mutburden_by_age)$r.squared
confint(lm.mean_mutburden_by_age)
  
sum_of_vaf_plot_log<-sum_of_vaf_df%>%
  mutate(exp_ID=gsub("8 pcw","8pcw",exp_ID))%>%
  mutate(sum_of_vaf=ifelse(sum_of_vaf==0,0.001,sum_of_vaf))%>%
  ggplot(aes(x=forcats::fct_reorder(Sample,sum_of_vaf),y=sum_of_vaf))+
  geom_point(size=0.1)+
  theme_classic()+
  my_theme+
  theme(axis.text.x=element_blank(),axis.ticks.x=element_blank(),strip.text.x = element_text(size=6))+
  facet_grid(cols=vars(factor(exp_ID,levels=ref_df$Sample[order(ref_df$Age)])),scales = "free",space="free")+
  scale_y_log10(breaks=c(0.005,0.01,0.03,0.06,0.12,0.25,0.5,1,2,4,8))+
  geom_hline(aes(yintercept=median),linetype=2,data=sum_of_vaf_summary%>%mutate(median=ifelse(median==0,0.001,median)),col="red",size=0.5)+
  geom_text(aes(x=0,y=median_pos,label=paste0("tilde(x) == ",round(median,3))),size=2,nudge_x=+115,nudge_y=+0.5,data=sum_of_vaf_summary%>%mutate(median_pos=ifelse(median==0,0.001,median)),parse=T)+
  labs(x="Sample",y="Mutation burden\n(sum of VAF)")+
  theme(panel.spacing.x=unit(1, "mm"))

sum_of_vaf_plot_log

ggsave(filename = paste0(figures_dir,"Figure_01/sum_of_vaf_log.pdf"),plot=sum_of_vaf_plot_log,height=2,width=7)
```

Now coloured by cell type
```{r}
sum_of_vaf_plot_log_by_celltype<-sum_of_vaf_df%>%
  mutate(exp_ID=gsub("8 pcw","8pcw",exp_ID))%>%
  mutate(sum_of_vaf=ifelse(sum_of_vaf==0,0.001,sum_of_vaf))%>%
  ggplot(aes(x=forcats::fct_reorder(Sample,sum_of_vaf),y=sum_of_vaf))+
  geom_point(aes(col=Cell_type),size=0.1,alpha=0.6)+
  scale_color_manual(values=c("#1a80bb", "#ea801c"))+
  theme_classic()+
  my_theme+
  theme(axis.text.x=element_blank(),axis.ticks.x=element_blank(),strip.text.x = element_text(size=6))+
  facet_grid(cols=vars(factor(exp_ID,levels=ref_df$Sample[order(ref_df$Age)])),scales = "free",space="free")+
  scale_y_log10(breaks=c(0.005,0.01,0.03,0.06,0.12,0.25,0.5,1,2,4,8))+
  geom_hline(aes(yintercept=median),linetype=2,data=sum_of_vaf_summary%>%mutate(median=ifelse(median==0,0.001,median)),col="red",size=0.5)+
  geom_text(aes(x=0,y=median_pos,label=paste0("tilde(x) == ",round(median,3))),size=2,nudge_x=+115,nudge_y=+0.5,data=sum_of_vaf_summary%>%mutate(median_pos=ifelse(median==0,0.001,median)),parse=T)+
  labs(x="Sample",y="Mutation burden\n(sum of VAF)",col="Cell type")+
  guides(color=guide_legend(override.aes = list(size=0.4,alpha=1)))+
  theme(panel.spacing.x=unit(1, "mm"),legend.position = "bottom")

sum_of_vaf_plot_log_by_celltype

ggsave(filename = paste0(figures_dir,"Figure_01/sum_of_vaf_log_by_celltype.pdf"),plot=sum_of_vaf_plot_log_by_celltype,height=2.2,width=7)
```

Comparison of mutation burden by cell type - violin plot, with overlaid comparisons using the non-parametric Wilcoxon test.
```{r}
individuals_with_both=c("18pcw","CB002","SX001","AX001","KX004")
sum_of_vaf_by_celltype<-sum_of_vaf_df%>%
  filter(exp_ID%in%individuals_with_both)%>%
  ggplot(aes(x=Cell_type,y=sum_of_vaf))+
  geom_violin(aes(fill=Cell_type),alpha=0.2)+
  scale_color_manual(values=c("#1a80bb", "#ea801c"))+
  scale_fill_manual(values=c("#1a80bb", "#ea801c"))+
  ggpubr::stat_compare_means(method = "wilcox.test",label = "p.format",fontface="italic",vjust = +1,size=2.6)+
  geom_jitter(aes(col=Cell_type),width=0.2,alpha=0.5)+
  facet_grid(cols=vars(factor(exp_ID,levels=ref_df$Sample[order(ref_df$Age)])),scales = "free",space="free")+
  theme_classic()+
  my_theme+
  labs(y="Mutation burden\n(sum of VAF)")+
  theme(axis.title.x=element_blank(),legend.position = "none")

sum_of_vaf_by_celltype

ggsave(filename = paste0(rebuttal_figs_dir,"sum_of_vaf_by_celltype_comparison.pdf"),plot=sum_of_vaf_by_celltype,height=2.2,width=7)

```


## MUTATIONAL PROFILES

Look at the 96-profile of the mutations, divied by mutation type
In fact - double this to a 192-profile, incorporating - in addition - the replication strand (heavy vs light).

```{r}
# read in the trinucleotide context reference object
mtdna_trinuc_freq <- readRDS(paste0(root_dir,"/data/mtDNA_trinuc_freqs_coding_dloop_heavy_light.Rds"))

# set regions for coding and dloop regions
coding_region <- 577:16023
d_loop_region <- c(1:576,16024:16569)
genomeFile="~/R_work/reference_files/genome.fa"

head(mtdna_trinuc_freq)
```

Plot all the mutations across samples.
```{r}
trinucleotide_plot(mutations = df_tidy%>%
                     filter(Sample!="global"&
                              #ML_Sig=="N1"&
                              !mut_ref%in%exclude_muts)%>%
                     separate(mut_ref,into=c("chr","pos","ref","mut"))%>%
                     dplyr::rename("donor"=exp_ID)%>%
                     mutate(pos=as.numeric(pos)),
                   analysis_region = "all_mtDNA",
                   analysis_type = "obs_exp",
                   file_name = paste0(plots_dir,"Mutational_signature.pdf"))
```

Now limit to check the 0.5 - 1% VAF mutations.
```{r}
trinucleotide_plot(mutations = df_tidy%>%
                     filter(Sample!="global"&
                              #ML_Sig=="N1"&
                              !mut_ref%in%exclude_muts&
                              vaf<0.01 & vaf>0.005)%>%
                     separate(mut_ref,into=c("chr","pos","ref","mut"))%>%
                     dplyr::rename("donor"=exp_ID)%>%
                     mutate(pos=as.numeric(pos)),
                   analysis_region = "all_mtDNA",
                   analysis_type = "obs_exp",
                   file_name = paste0(plots_dir,"Mutational_signature_0.005_0.01.pdf"))
```

Now check the <0.5 VAF mutations.
```{r}
trinucleotide_plot(mutations = df_tidy%>%
                     filter(Sample!="global"&
                              ML_Sig=="N1"&
                              !mut_ref%in%exclude_muts&
                              vaf<0.005)%>%
                     separate(mut_ref,into=c("chr","pos","ref","mut"))%>%
                     dplyr::rename("donor"=exp_ID)%>%
                     mutate(pos=as.numeric(pos)),
                   analysis_region = "all_mtDNA",
                   analysis_type = "obs_exp",
                   file_name = paste0(plots_dir,"Mutational_signature_<0.005.pdf"))

```

## ANALYSIS OF MUTATION BURDEN ACCORDING TO CLONE EXPANSION/ SINGLETON

Do clonal expansions have higher mutation burdens compared to singleton colonies?
Need to take into account that different cells within a clade are not independent, therefore take an average
This may slightly alter the distribution of mutation burdens as some of the variation is smoothed out (might effect the k-s test statistic)
Therefore, could alternatively randomly pick a single sample.

```{r}
exp_vs_singleton_sov<-Map(list=mito_data[3:10],exp_ID=names(mito_data)[3:10],function(list,exp_ID) {
  
  #Using the 'get_expanded_clade_nodes' function to effectively cut across the tree at 100 mutations of molecular time
  #Some of the clades will be singletons
  ec_df<-get_expanded_clade_nodes(list$tree.ultra,height_cut_off = 100,min_clonal_fraction = 0,min_samples=1)
  
  ec_df$mean_sov<-sapply(ec_df$nodes,function(node) {
    Samples<-getTips(list$tree.ultra,node=node)
    mean_sov_in_clade<-sum_of_vaf_df%>%filter(Sample%in%Samples)%>%pull(sum_of_vaf)%>%mean(na.rm=T)
    return(mean_sov_in_clade)
  })
  
  return(ec_df%>%mutate(exp_ID=exp_ID,.before=1))
  
})%>%dplyr::bind_rows()

exp_vs_singleton_sov%>%
  mutate(exp_ID=factor(exp_ID,levels=ref_df%>%filter(Age>0)%>%arrange(Age)%>%pull(Sample)))%>%
  mutate(type=ifelse(n_samples==1,"Sing.","Exp."))%>%
  ggplot(aes(x=type,y=mean_sov,col=type,fill=type))+
  geom_violin(alpha=0.1)+
  geom_jitter(width=0.1,height=0,alpha=0.3)+
  scale_color_brewer(palette="Set1")+
  scale_fill_brewer(palette="Set1")+
  facet_grid(~exp_ID)+
  theme_classic()+
  theme(legend.position="none")+
  ggpubr::stat_compare_means(method = "wilcox.test",label = "p.format",fontface="italic")+
  labs(x="",y="Mean mtDNA mutation burden\n (sum of VAF)")

all.ks.tests<-lapply(unique(exp_vs_singleton_sov$exp_ID),function(this_exp_ID) {
  ks<-ks.test(x=exp_vs_singleton_sov%>%filter(n_samples==1 & exp_ID==this_exp_ID)%>%pull(mean_sov),
          y=exp_vs_singleton_sov%>%filter(n_samples>1 & exp_ID==this_exp_ID)%>%pull(mean_sov))
  return(ks)
})

p.adjust(sapply(all.ks.tests,function(res) res$p.value),method = "BH")

```

## Mutation burden by driver status: mut/ wt

Step 1. Add basic nDNA driver mutation info to the object
```{r}
mito_data<-Map(list=mito_data,exp_ID=names(mito_data),function(list,exp_ID) {
  annot_muts_file<-paste0(root_dir,"/data/blood_adult/annot_files_filtered/annotated_muts_filt_",exp_ID,".Rds")
  if(file.exists(annot_muts_file)) {
    cat(paste("Importing annotated mutations dataframe for",exp_ID),sep="\n")
    list$nDNA_mats<-readRDS(annot_muts_file)
    
    if(nrow(list$nDNA_mats$mat)==0) {
      stop(return(list))
    }
    
    tree_samples<-list$tree.ultra$tip.label[-which(list$tree.ultra$tip.label=="Ancestral")]
    
    if(any(!tree_samples%in%colnames(list$nDNA_mats$NV))) {
      cat("Dropping samples not found in the count matrix.")
      drop_tips<-tree_samples[!tree_samples%in%colnames(list$nDNA_mats$NV)]
      list$tree.ultra<-drop.tip(list$tree.ultra,tip = drop_tips)
      list$tree<-drop.tip(list$tree,tip = drop_tips)
      tree_samples<-tree_samples[tree_samples%in%colnames(list$nDNA_mats$NV)]
    }
    
    mtr<-as.matrix(cbind(list$nDNA_mats$NV[,tree_samples,drop=F],matrix(0,ncol = 1,nrow = nrow(list$nDNA_mats$NV),dimnames = list(rownames(list$nDNA_mats$NV),"Ancestral"))))
    dep<-as.matrix(cbind(list$nDNA_mats$NR[,tree_samples,drop=F],matrix(10,ncol = 1,nrow = nrow(list$nDNA_mats$NR),dimnames = list(rownames(list$nDNA_mats$NV),"Ancestral"))))
    res<-treemut::assign_to_tree(tree=list$tree.ultra,mtr=mtr,dep=dep)
    list$nDNA_mats$mat$node<-res$tree$edge[res$summary$edge_ml,2]
    return(list)
  } else {
    return(list)
  }
})
```

Step 2. Generate a similar 'sum of vaf' dataframe, but now incorporating driver information

```{r mutburden_by_driverstatus,fig.height=2,fig.width=6}
sov_comparison_df<-Map(list=mito_data[3:10],exp_ID=names(mito_data)[3:10],function(list,exp_ID) {
  cat(exp_ID,sep="\n")
  #Using the 'get_expanded_clade_nodes' function to effectively cut across the tree at 100 mutations of molecular time
  #Some of the clades will be singletons
  ec_df<-get_expanded_clade_nodes(list$tree.ultra,height_cut_off = 100,min_clonal_fraction = 0,min_samples=1)
  
  ec_df$mean_sov<-sapply(ec_df$nodes,function(node) {
    Samples<-getTips(list$tree.ultra,node=node)
    mean_sov_in_clade<-sum_of_vaf_df%>%filter(Sample%in%Samples)%>%pull(sum_of_vaf)%>%mean(na.rm=T)
    return(mean_sov_in_clade)
  })
  
  left_join(ec_df,list$nDNA_mats$mat%>%dplyr::select(node,mut_ref,Gene,CDS,Protein),by=c("nodes"="node"))%>%
    mutate(exp_ID=exp_ID,.before=1)
  
})%>%dplyr::bind_rows()

# Split samples by whether they harbour a DTA driver mutation or not
mut_set=c("DNMT3A","TET2","ASXL1")
IDs_to_include=paste0("KX00",c(7,8,4,3))

sov_mut_vs_wt_df<-sov_comparison_df%>%
  mutate(exp_ID=factor(exp_ID,levels=ref_df%>%filter(Age>0)%>%arrange(Age)%>%pull(Sample)))%>%
  replace_na(list(Gene=""))%>%
  mutate(type=ifelse(Gene%in%mut_set,"mut","wt"))%>%
  filter(exp_ID%in%IDs_to_include)

sov_mut_vs_wt<-sov_mut_vs_wt_df%>%
  ggplot(aes(x=type,y=mean_sov,col=type,fill=type))+
  geom_violin(alpha=0.1)+
  geom_jitter(width=0.1,height=0,alpha=0.3)+
  scale_color_brewer(palette="Set1")+
  scale_fill_brewer(palette="Set1")+
  facet_grid(~exp_ID)+
  theme_classic()+
  my_theme+
  theme(legend.position="none")+
  ggpubr::stat_compare_means(method = "wilcox.test",label = "p.format",fontface="italic",vjust = +0.5,size=3)+
  labs(x="",y="Mean mtDNA mutation burden\n (sum of VAF)")

sov_mut_vs_wt

ggsave(filename = paste0(rebuttal_figs_dir,"sov_mut_vs_wt.pdf"),sov_mut_vs_wt,width=3.3,height=2.5)
```

Do a formal regression using a linear mixed effects model: does the mutation burden differ by driver status?  Individual is included here as a random effect.

```{r}
mut_vs_wt.lmer<-lmerTest::lmer(mean_sov~type+(1|exp_ID),data=sov_mut_vs_wt_df)
summary(mut_vs_wt.lmer)
confint(mut_vs_wt.lmer)

```

## Mutation burden by clone status: singleton vs part of clone

Assign cells as to whether they are unrelated to any other cell after development (defined here as after 100 mutations of molecular time, although the exact cut off has little impact on results as long at is is ~50 - 150).
Now "mut_samples" means that they are part of a clonal expansion (≥2 cells related after development)
"wt_samples" means that they are singletons.

```{r mutburden_by_clonestatus,fig.height=2,fig.width=6}
mut_vs_wt_list<-Map(list=mito_data[5:8],exp_ID=names(mito_data)[5:8],function(list,exp_ID) {
  cat(exp_ID,sep="\n")
  #Using the 'get_expanded_clade_nodes' function to effectively cut across the tree at 100 mutations of molecular time
  #Some of the clades will be singletons
  ec_df<-get_expanded_clade_nodes(list$tree.ultra,height_cut_off = 100,min_clonal_fraction = 0,min_samples=1)%>%
    left_join(list$nDNA_mats$mat%>%dplyr::select(node,mut_ref,Gene,CDS,Protein),by=c("nodes"="node"))%>%
    mutate(exp_ID=exp_ID,.before=1)
  
  mut_nodes<-ec_df%>%filter(n_samples>1)%>%pull(nodes)
  wt_nodes<-ec_df%>%filter(n_samples==1)%>%pull(nodes)

  mut_samples<-unlist(lapply(mut_nodes,function(node) sample(x=getTips(list$tree.ultra,node=node),size=1)))
  wt_samples<-unlist(lapply(wt_nodes,function(node) sample(x=getTips(list$tree.ultra,node=node),size=1)))
  
  return(list(mut=mut_samples,wt=wt_samples))
})

all_mut<-unlist(lapply(mut_vs_wt_list,function(x) x$mut))
all_wt<-unlist(lapply(mut_vs_wt_list,function(x) x$wt))

IDs_to_include=paste0("KX00",c(7,8,4,3))

exp_vs_singleton_df<-sov_comparison_df%>%
  mutate(exp_ID=factor(exp_ID,levels=ref_df%>%filter(Age>0)%>%arrange(Age)%>%pull(Sample)))%>%
  filter(exp_ID%in%IDs_to_include)%>%
  mutate(type=ifelse(n_samples>1,"Clonal\nexpansion",ifelse(n_samples==1,"Singleton",NA)))%>%
  filter(!is.na(type))

sov_exp_vs_singleton<-exp_vs_singleton_df%>%
  ggplot(aes(x=type,y=mean_sov,col=type,fill=type))+
  geom_violin(alpha=0.1)+
  geom_jitter(width=0.1,height=0,alpha=0.3)+
  scale_color_brewer(palette="Set1")+
  scale_fill_brewer(palette="Set1")+
  facet_grid(~exp_ID)+
  theme_classic()+
  my_theme+
  theme(legend.position="none",axis.text.x=element_text(angle=0))+
  ggpubr::stat_compare_means(method = "wilcox.test",label = "p.format",fontface="italic",vjust = +0.5,size=3)+
  labs(x="",y="Mean mtDNA mutation burden\n (sum of VAF)")

sov_exp_vs_singleton

ggsave(filename = paste0(rebuttal_figs_dir,"sov_exp_vs_singleton.pdf"),sov_exp_vs_singleton,width=3.3,height=2.5)
```

Again, do a formal regression with a linear mixed effects regression model.

```{r}
exp_vs_singleton_df%>%group_by(type)%>%dplyr::summarise(n=n())
exp_vs_singleton.lmer<-lmerTest::lmer(mean_sov~type+(1|exp_ID),data=exp_vs_singleton_df)
summary(exp_vs_singleton.lmer)
confint(exp_vs_singleton.lmer)

```

## SELECTION ANALYSIS

This section uses two approaches to look for evidence of selection for or against mtDNA mutations.

(1) Cumulative distribution function of heteroplasmy levels, divided by the mutation impact (i.e. synonymous, missense or nonsense).
(2) dN/dS. This is the ratio of non-synonymous mutations compared to synonymous, corrected such that the anticipated ratio by chance is 1.  A ratio of >1 implies overall positive selection (i.e. there is an excess of non-synonymous variants), a ratio of <1 implies overall negative selection (i.e. there is a depletion of non-synonymous variants), and a ratio of ~1 implies overall neutrality.

```{r annotate_variants_with_dndscv}
library(dndscv)
mtref_rda_path=ifelse(Sys.info()['sysname']=="Darwin",paste0(root_dir,"/data/mtref.rda"),"/lustre/scratch126/casm/team154pc/ms56/Mitochondria_study/mtref.rda")
mtref_rda_path="~/Downloads/mtref.rda"
input.dir <- paste0(root_dir,"dnds_tables/")

dnds_theme<-theme(panel.border = element_rect(color = "black",
                                              fill = NA,
                                              linewidth = 0.75),
                  strip.text = element_text(face="plain", size=6, colour = "black",),
                  strip.background = element_rect(fill="white", colour="black", linewidth =1),
                  axis.text.x = element_text(color = "black", size = 6, angle = 0, hjust = .5, vjust = 0.5, face = "plain"),
                  axis.text.y = element_text(color = "black", size = 6, angle = 0, hjust = 1, vjust = 0, face = "plain"),  
                  axis.title.x = element_text(color = "black", size = 8, angle = 0, hjust = .5, vjust = 0, face = "plain"),
                  axis.title.y = element_text(color = "black", size = 8, angle = 90, hjust = .5, vjust = .5, face = "plain", 
                                              margin = margin(t = 0, r = 10, b = 0, l = 0)), 
                  plot.title = element_text(color = "black", size = 10, hjust = .5, face = "plain"), 
                  legend.position = "right")

#exclude ND6 from analysis since its on the other strand
all_mtDNA_genes <- c("MT-CYB", "MT-ND5", "MT-ND2", "MT-ND4", "MT-ND1", "MT-CO3", "MT-ATP6","MT-ND3", "MT-ATP8", "MT-ND4L", "MT-CO2", "MT-CO1","MT-ND5","MT-ND6")
target_genes <- c("MT-CYB", "MT-ND5", "MT-ND2", "MT-ND4", "MT-ND1", "MT-CO3", "MT-ATP6","MT-ND3", "MT-ATP8", "MT-ND4L", "MT-CO2", "MT-CO1")

# This labels each sample as foetal, cord blood or adult
df_tidy$Tissue<-ifelse(grepl("pcw",df_tidy$exp_ID),"Foetal blood",ifelse(grepl("CB",df_tidy$exp_ID),"Cord blood","Adult blood"))
valid.tissues <- unique(df_tidy$Tissue)

df_tidy_annotated<-lapply("x",function(tissue.id) {
  cat(tissue.id,sep="\n")
  
  # read in the mtdna variant file and remove patient id
  mtdna.variant.data <- df_tidy%>%
    #dplyr::filter(Tissue==tissue.id)%>%
    separate("mut_ref", c("chr", "pos", "ref", "mut"), "_")%>%
    mutate(pos=as.numeric(pos))%>%
    dplyr::select("sampleID"=Sample,chr,pos,ref,mut,vaf)%>% # next, rename/reorder the columns to make them compatible with dnds input format
    dplyr::filter(!duplicated(.))%>% # remove duplicated variants
    arrange(sampleID,pos)
  
  # run dnds with the mtDNA variants to annotate them (the selection analysis isn't actually used here)
  mtdna.dndsout <- dndscv(mtdna.variant.data,
                          gene_list=all_mtDNA_genes, 
                          refdb = mtref_rda_path,
                          numcode = 2,
                          max_coding_muts_per_sample = Inf,
                          max_muts_per_gene_per_sample = Inf)
  
  # get the results with the annotated variants
  annotated.mtdna.variants <- left_join(mtdna.variant.data,mtdna.dndsout$annotmuts,by=c("sampleID","chr","pos","ref","mut"))%>%
    tidyr::replace_na(replace=list(impact="Non-Coding"))%>%
    mutate(tissue=tissue.id,.before=1)
  return(annotated.mtdna.variants)
})%>%dplyr::bind_rows()

complete.annotated.mutation.table<-df_tidy_annotated%>%
  left_join(df_tidy%>%dplyr::bind_rows()%>%dplyr::select(Sample,exp_ID),by=c("sampleID"="Sample"),relationship="many-to-many")%>%
  dplyr::rename("patientID"=exp_ID)%>%
  filter(!duplicated(.))%>%
  arrange(patientID,pos)%>% # order the dataframe
  tidyr::unite(col="mut_ref",chr,pos,ref,mut,sep="_",remove=F)

head(complete.annotated.mutation.table)
```

### Compare annotation with nuclear vs mitochondrial genetic code.

Run analysis with the (incorrect) nuclear annotation for comparison.
This is because in an initial version of the analysis, we used the incorrect genetic code, so want to show here how this initially altered the results, and how it compares with the correct annotation.

```{r nuclear_genetic_code}

df_tidy_annotated_NUC<-lapply("x",function(tissue.id) {
  cat(tissue.id,sep="\n")
  
  # read in the mtdna variant file and remove patient id
  mtdna.variant.data <- df_tidy%>%
    #dplyr::filter(Tissue==tissue.id)%>%
    separate("mut_ref", c("chr", "pos", "ref", "mut"), "_")%>%
    mutate(pos=as.numeric(pos))%>%
    dplyr::select("sampleID"=Sample,chr,pos,ref,mut,vaf)%>% # next, rename/reorder the columns to make them compatible with dnds input format
    dplyr::filter(!duplicated(.))%>% # remove duplicated variants
    arrange(sampleID,pos)
  
  # run dnds with the mtDNA variants to annotate them (the selection analysis isn't actually used here)
  mtdna.dndsout <- dndscv(mtdna.variant.data,
                          gene_list=all_mtDNA_genes, 
                          refdb = mtref_rda_path,
                          numcode = 1,
                          max_coding_muts_per_sample = Inf,
                          max_muts_per_gene_per_sample = Inf)
  
  # get the results with the annotated variants
  annotated.mtdna.variants <- left_join(mtdna.variant.data,mtdna.dndsout$annotmuts,by=c("sampleID","chr","pos","ref","mut"))%>%
    tidyr::replace_na(replace=list(impact="Non-Coding"))%>%
    mutate(tissue=tissue.id,.before=1)
  return(annotated.mtdna.variants)
})%>%dplyr::bind_rows()

complete.annotated.mutation.table.NUC<-df_tidy_annotated_NUC%>%
  left_join(df_tidy%>%dplyr::bind_rows()%>%dplyr::select(Sample,exp_ID),by=c("sampleID"="Sample"),relationship="many-to-many")%>%
  dplyr::rename("patientID"=exp_ID)%>%
  filter(!duplicated(.))%>%
  arrange(patientID,pos)%>% # order the dataframe
  #mutate(impact=ifelse(impact%in%c("Stop_loss","Nonsense"),"Truncating",impact))%>% #rename stop loss and nonsense mutations
  tidyr::unite(col="mut_ref",chr,pos,ref,mut,sep="_",remove=F)

```

```{r consequence_change_plot}
## Plots of nuclear vs mitochondrial annotation----
impact_comparison_df<-left_join(complete.annotated.mutation.table,
                                complete.annotated.mutation.table.NUC%>%dplyr::select(tissue,sampleID,mut_ref,codonsub,"impact_NUC"=impact))%>%
  dplyr::select(tissue,sampleID,mut_ref,codonsub,impact,impact_NUC)

consequence_change_plot<-impact_comparison_df%>%
  mutate(change=paste(impact_NUC,impact,sep=">"))%>%
  mutate(altered=ifelse(impact_NUC==impact,"unchanged","changed"))%>%
  filter(impact!="Non-Coding")%>%
  mutate(altered=factor(altered,levels=c("unchanged","changed")))%>%
  ggplot(aes(y=change,fill=altered))+
  geom_bar()+
  facet_grid(rows=vars(altered),scales="free",space="free_y")+
  theme_classic()+
  scale_fill_manual(values=rev(c("#e06010","#2d6e3e")),guide="none")+
  my_theme+
  labs(y="Consequence of updating to\nmitochondrial genetic code")

consequence_change_plot

ggsave(filename = paste0(plots_dir,"consequence_change_plot.pdf"),consequence_change_plot,width=3.5,height=2.5)
```


```{r}
impact_comparison_df%>%
  filter(impact!="Non-Coding")%>%
  mutate(change=paste(impact_NUC,impact,sep=">"))%>%
  mutate(altered=ifelse(impact_NUC==impact,"unchanged","changed"))%>%
  pull(altered)%>%table()
```


```{r}
total_numbers_old_vs_new_plot<-impact_comparison_df%>%
  dplyr::select(mut_ref,"corrected_annotation"=impact,"original_annotation"=impact_NUC)%>%
  tidyr::gather(-mut_ref,key="annotation",value="impact")%>%
  filter(impact!="Non-Coding")%>%
  mutate(impact=factor(impact,levels=c("Synonymous","Missense","Nonsense","Stop_loss")))%>%
  ggplot(aes(y = impact,fill=annotation))+
  geom_bar(position="dodge",width=0.6,col="black",linewidth=0.1)+
  scale_fill_brewer(palette="Set2")+
  theme_classic()+
  my_theme+
  labs(fill="",y="")+
  theme(legend.position = "top")

total_numbers_old_vs_new_plot

ggsave(filename = paste0(plots_dir,"total_numbers_old_vs_new_plot.pdf"),total_numbers_old_vs_new_plot,width=3,height=2.5)
```


```{r}
impact_comparison_df%>%
  dplyr::select(mut_ref,"new_annotation"=impact,"original_annotation"=impact_NUC)%>%
  tidyr::gather(-mut_ref,key="annotation",value="impact")%>%
  filter(impact!="Non-Coding")%>%
  dplyr::group_by(annotation,impact)%>%
  dplyr::summarise(n=n())
```


```{r}
simple_impact_comparison_df<-impact_comparison_df%>%
  mutate(change=paste(impact_NUC,impact,sep=">"))%>%
  mutate(altered=ifelse(impact_NUC==impact,"unchanged","changed"))

new_muts<-simple_impact_comparison_df%>%
  filter(altered=="changed")%>%
  group_by(impact)%>%
  dplyr::summarise(new=n())

lost_muts<-simple_impact_comparison_df%>%
  filter(altered=="changed")%>%
  group_by(impact_NUC)%>%
  dplyr::summarise(lost=n())%>%dplyr::rename("impact"=impact_NUC)

simple_impact_comparison_df%>%
  filter(altered=="unchanged" & impact!="Non-Coding")%>%
  group_by(impact)%>%
  dplyr::summarise(unchanged=n())%>%
  left_join(new_muts)%>%left_join(lost_muts)%>%
  tidyr::gather(-impact,key="result",value="n")%>%
  ggplot(aes(y=impact,x=n,fill=result,alpha = result!="lost"))+
  geom_bar(stat="identity")+
  scale_alpha(range=c(0.4,1),guide="none")+
  theme_classic()+
  my_theme+
  theme(axis.title.y=element_blank(),legend.title = element_blank())

```

### Look at the numbers of synonymous/ non-synonymous mutations

```{r}
my_comparisons <- list(c("Missense", "Synonymous"),
                       c("Synonymous", "Nonsense"))

mutation_category_by_vaf_violin_plot<-complete.annotated.mutation.table%>%
  dplyr::filter(impact!="Non-Coding"&impact!="Stop_loss")%>%
  ggplot(aes(impact, vaf, color = impact)) +
  geom_violin(aes(fill=impact),alpha=0.2) +
  geom_jitter(width=0.25,size = .1, alpha = 0.1) +
  theme_classic() + 
  scale_color_manual(values=mut_type_cols) +
  scale_fill_manual(values=mut_type_cols) +
  scale_y_log10() + 
  #facet_wrap(~ tissue, scales='free_x', nrow = 1) +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) + 
  ggpubr::stat_compare_means(comparisons = my_comparisons, size = 2) +
  labs(x="",y="Variant allele fraction",col="Impact",fill="Impact") + 
  my_theme+
  theme(legend.position="none")
```


```{r}
#Gives some sense of differing heteroplasmy distributions
#But bear in mind there are multiple non-independent measures for some variants
#Would need to account for this e.g. by taking single measure of e.g. max VAF of variant per individual
syn_vals <- complete.annotated.mutation.table %>%
  dplyr::filter(impact == "Synonymous") %>% pull(vaf)

mis_vals <- complete.annotated.mutation.table %>%
  dplyr::filter(impact == "Missense") %>% pull(vaf)

nons_vals <- complete.annotated.mutation.table %>%
  dplyr::filter(impact == "Nonsense") %>% pull(vaf)

ks_syn_mis <- ks.test(syn_vals, mis_vals)
ks_syn_non <- ks.test(syn_vals, nons_vals)

find_ks_coords <- function(vals1, vals2) {
  ecdf1 <- ecdf(vals1)
  ecdf2 <- ecdf(vals2)
  all_y <- sort(c(vals1, vals2))
  diffs <- abs(ecdf1(all_y) - ecdf2(all_y))
  max_y <- all_y[which.max(diffs)]
  data.frame(
    y    = max_y,
    xmin = min(ecdf1(max_y), ecdf2(max_y)),
    xmax = max(ecdf1(max_y), ecdf2(max_y))
  )
}

fmt_p <- function(p) {
  ifelse(p < 0.001, "p < 0.001",
         ifelse(p < 0.01,  paste0("p = ", round(p, 3)),
                paste0("p = ", round(p, 2))))
}

coords_non <- find_ks_coords(syn_vals, nons_vals)
coords_mis <- find_ks_coords(syn_vals, mis_vals)

sapply(list(mis_vals,nons_vals,syn_vals),length)

mutation_type_ecdf_plot<-complete.annotated.mutation.table %>%
  filter(impact != "Non-Coding" & impact != "Stop_loss") %>%
  ggplot(aes(y = vaf, col = impact)) +
  stat_ecdf(geom = "step", linewidth = 0.8) +
  # Syn vs Missense segment
  annotate("segment",
           x = coords_mis$xmin, xend = coords_mis$xmax,
           y = coords_mis$y,    yend = coords_mis$y,
           colour = "black", linewidth = 0.5, linetype = "dashed") +
  annotate("text",
           x = mean(c(coords_mis$xmin, coords_mis$xmax)),
           y = coords_mis$y,
           vjust = -0.5, size = 2,
           label = paste0("Syn vs Mis\n", fmt_p(ks_syn_mis$p.value))) +
  # Syn vs Nonsense segment
  annotate("segment",
           x = coords_non$xmin, xend = coords_non$xmax,
           y = coords_non$y,    yend = coords_non$y,
           colour = "black", linewidth = 0.5, linetype = "dashed") +
  annotate("text",
           x = mean(c(coords_non$xmin, coords_non$xmax)),
           y = coords_non$y,
           vjust = -0.5, size = 2,
           label = paste0("Syn vs Nons\n", fmt_p(ks_syn_non$p.value)))+
  scale_y_continuous(limits = c(0.01, 1),
                     labels = scales::percent_format(accuracy = 1),
                     name = "Variant Allele Fraction (VAF)") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1),
                     name = "Cumulative fraction of variants") +
  theme_classic(base_size = 13) +
  scale_color_manual(values = mut_type_cols) +
  my_theme +
  theme(panel.grid.minor = element_blank(),
        legend.position = c(0.3, 0.8)) +
  labs(col = "Mutation type")

mutation_type_ecdf_plot

ggsave(filename = paste0(rebuttal_figs_dir,"mutation_type_ecdf_plot.pdf"),mutation_type_ecdf_plot,width=2.5,height=2.5)
```


```{r mutation_type_by_gene,fig.height=2.5,fig.width=3}

mutation_type_by_gene<-complete.annotated.mutation.table%>%
  filter(impact!="Non-Coding")%>%
  distinct(sampleID,mut_ref,gene,impact)%>%
  group_by(gene,impact)%>%
  dplyr::summarise(n=n())%>%
  ggplot(aes(y=gene,x=n,fill=impact))+
  geom_bar(position="stack",stat="identity",col="black",linewidth=0.1)+
  scale_fill_manual(values=mut_type_cols)+
  scale_x_continuous(breaks=seq(0,1400,200))+
  theme_classic()+
  my_theme+
  theme(axis.title.y=element_blank(),
        axis.text.y=element_text(face="italic"))+
  labs(x="Number of variants",fill="Mutation\nconsequence")

mutation_type_by_gene

ggsave(filename = paste0(rebuttal_figs_dir,"mutation_type_by_gene_blood.pdf"),mutation_type_by_gene,width=3.5,height=2.5)

```

### dndscv across the blood individuals

Combine across all the blood individuals and assess the dN/dS ratios of all variants in aggregate.
Only count each variant once per individual.
As before, using the argument "numcode=2" ensures that the mitochondrial genetic code is used for annotation.

```{r}
dndscv_by_tissue<-lapply("x",function(tissue.id) {
  
  cat(tissue.id,sep="\n")
  tissue_info<-df_tidy%>%
    separate("mut_ref", c("chr", "pos", "ref", "mut"), "_")%>%
    mutate(pos=as.numeric(pos))%>%
    dplyr::select("sampleID"=exp_ID,chr,pos,ref,mut)%>%
    dplyr::filter(!duplicated(.))%>%
    arrange(sampleID,pos)
  
  mtdna.dndsout <- dndscv(tissue_info,
                          gene_list=target_genes, 
                          refdb = mtref_rda_path,
                          numcode = 2,
                          max_coding_muts_per_sample = Inf,
                          max_muts_per_gene_per_sample = Inf)
  return(mtdna.dndsout)
})

# processing the output to get a combined 'globaldnds' dataframe by tissue type
rename_vec=c("Missense","Nonsense","Truncating","Overall")
names(rename_vec)=c("wmis","wnon","wtru","wall")

globaldnds_res<-Map(dndsout=dndscv_by_tissue,tissue="x",function(dndsout,tissue) {
  dndsout$globaldnds%>%
    filter(!is.na(name) & complete.cases(.))%>%
    mutate(name=rename_vec[name])%>%
    mutate(tissue=tissue)
})%>%dplyr::bind_rows()

tissue_order=c("Foetal blood","Cord blood","Adult blood")
nb.cols <- length(tissue_order)
mycolors <- colorRampPalette(ggsci::pal_lancet(palette = "lanonc")(9))(nb.cols)
```

Visualize these results
```{r}
max_dnds_value<-1.3
stats_to_include=c("Missense","Truncating","Overall")
global_dnds_blood<-globaldnds_res%>%
  filter(name%in%stats_to_include)%>%
  mutate(cihigh=ifelse(cihigh>max_dnds_value,max_dnds_value,cihigh),
         tissue=factor(tissue,levels=tissue_order),
         name=factor(name,levels=stats_to_include))%>%
  ggplot(aes(x=name, y = mle, ymin = cilow, ymax = cihigh, color = name, shape = name)) +
  geom_linerange(position= position_dodge2(width=0.75), linewidth = 0.5, color="darkgrey") +
  geom_point(position=position_dodge2(width=0.75), size = 1.5) +
  labs(y = "Mitochondrial genome-wide dN/dS") +
  theme_classic() + 
  ylim(c(0.5,max_dnds_value))+
  scale_x_discrete(labels = function(x) str_wrap(x, width = 6))+
  scale_color_manual(values = mut_type_cols, name="Mutation type") +
  scale_shape_manual(values = c(15, 16, 17, 18), name = "Mutation type") + 
  geom_hline(yintercept=1, linetype='dashed', col = 'darkgrey', linewidth = 0.5) +
  my_theme+
  #facet_wrap(~tissue,nrow=1)+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_text(angle=45,vjust = +0.5),
        legend.title=element_blank(),
        legend.position="none")

ggsave(filename = paste0(rebuttal_figs_dir,"global_dnds_blood.pdf"),global_dnds_blood,width=1.3,height=2.5)
```


```{r}
### Generate Extended Data FIG. 2C ---------
gene_order=c(paste0("MT-ND",1:6),"MT-ND4L","MT-CYB",paste0("MT-CO",1:3),"MT-ATP6","MT-ATP8")
gene_order<-gene_order[gene_order%in%target_genes]

selcv_res_by_tissue<-Map(dndsout=dndscv_by_tissue,tissue="x",function(dndsout,tissue) {
  temp<-dndsout$sel_cv%>%
    dplyr::select(gene_name,wmis_cv,wnon_cv,qmis_cv,qtrunc_cv)
  
  dplyr::bind_rows(temp%>%dplyr::select(gene_name,"dNdS"=wmis_cv,"qval"=qmis_cv)%>%mutate(type="Missense"),
                   temp%>%dplyr::select(gene_name,"dNdS"=wnon_cv,"qval"=qtrunc_cv)%>%mutate(type="Nonsense"))%>%
    mutate(tissue=tissue)
})%>%dplyr::bind_rows()

qval_cutoff=0.05
dnds_heatmap_by_gene<-selcv_res_by_tissue%>%
  mutate(dnds_if_signif=ifelse(qval<qval_cutoff,paste0(round(dNdS,2),"*"),""),
         gene_name=factor(gene_name,levels=rev(gene_order)))%>%
  ggplot(aes(y=gene_name,x=type,fill=dNdS,label=dnds_if_signif))+
  geom_tile()+
  geom_text(aes(col=type),size=1.6)+
  scale_color_manual(values=c("black","white"),guide="none")+
  theme_classic()+
  my_theme+
  theme(axis.text.x=element_text(angle=45,vjust=0.5),legend.position="none")+
  scale_fill_gradientn(colors=RColorBrewer::brewer.pal(n=11,name = "RdYlGn"),limits=c(0,2))+
  labs(x="",y="")

ggsave(filename = paste0(rebuttal_figs_dir,"dnds_heatmap_by_gene.pdf"),dnds_heatmap_by_gene,width=1.5,height=2.5)

```

### dndscv - split by heteroplasmy

```{r}
#Define the VAF 'bins' that will be used for comparing VAF distributions
#Can set this on a log scale, or as decile bins
boundaries<-c(0,0.01,0.1,0.2,0.5,1)
new_VAF_groups=paste0(100*head(boundaries,-1),"-",100*tail(boundaries,-1),"%")
VAF_groups=data.frame(labels=new_VAF_groups,LL=head(boundaries,-1),UL=tail(boundaries,-1))

#Assign each observed VAF into its 'VAF group' bin
df_tidy$VAF_group=sapply(df_tidy$vaf,function(vaf) {
  if(vaf==0) {
    return("Absent")
  } else {
    return(VAF_groups$labels[VAF_groups$LL<vaf & VAF_groups$UL>=vaf])
  }
})


dndscv_by_vaf_blood<-lapply(new_VAF_groups,function(VAF_bin) {
  cat(VAF_bin,sep="\n")
  tissue_vaf_info<-df_tidy%>%
    dplyr::filter(VAF_group==VAF_bin)%>%
    separate("mut_ref", c("chr", "pos", "ref", "mut"), "_")%>%
    mutate(pos=as.numeric(pos))%>%
    dplyr::select("sampleID"=exp_ID,chr,pos,ref,mut)%>% 
    dplyr::filter(!duplicated(.))%>% #only count each mutation once per individual in each VAF group
    arrange(sampleID,pos)
  
  mtdna.dndsout <- dndscv(tissue_vaf_info, gene_list=target_genes, 
                          refdb = mtref_rda_path,
                          max_coding_muts_per_sample = Inf,
                          numcode = 2,
                          max_muts_per_gene_per_sample = Inf)
  return(mtdna.dndsout)
})

globaldnds_res_by_vaf<-Map(dndsout=dndscv_by_vaf_blood,VAF_bin=new_VAF_groups,function(dndsout,VAF_bin) {
  dndsout$globaldnds%>%
    filter(!is.na(name) & complete.cases(.))%>%
    mutate(name=rename_vec[name])%>%
    mutate(VAF_group=VAF_bin)
})%>%dplyr::bind_rows()

max_dnds_value<-1.5
stats_to_include=c("Overall")
global_dnds_by_vaf_plot<-globaldnds_res_by_vaf%>%
  filter(VAF_group%in%VAF_groups$labels[2:length(boundaries)] & name%in%stats_to_include)%>%
  mutate(cihigh=ifelse(cihigh>max_dnds_value,max_dnds_value,cihigh),
         VAF_group=factor(VAF_group,levels=VAF_groups$labels[2:length(boundaries)]))%>%
  ggplot(aes(x=VAF_group, y = mle, ymin = cilow, ymax = cihigh, color = name, shape = name)) +
  geom_linerange(position= position_dodge2(width=0.75), size = 0.5, color="darkgrey") +
  geom_point(position=position_dodge2(width=0.75), size = 1.5) +
  labs(y = "Mitochondrial genome-wide dN/dS", x = "Variant allele fraction") +
  theme_classic() + 
  ylim(c(0,max_dnds_value))+
  scale_color_manual(values = mut_type_cols, name="Mutation type") +
  scale_shape_manual(values = 17, name = "Mutation type") + 
  geom_hline(yintercept=1, linetype='dashed', col = 'darkgrey', size = 0.5) +
  theme_classic()+
  my_theme+
  #facet_wrap(~tissue,nrow=1)+
  theme(legend.position="none",axis.text.x=element_text(angle=90))

ggsave(filename = paste0(rebuttal_figs_dir,"global_dnds_by_vaf_blood.pdf"),global_dnds_by_vaf_plot,width=1.6,height=2.5)

max_dnds_value<-2
stats_to_include=c("Truncating","Missense")
global_dnds_bytype_by_vaf_plot<-globaldnds_res_by_vaf%>%
  filter(VAF_group%in%VAF_groups$labels[2:length(boundaries)] & name%in%stats_to_include)%>%
  mutate(cihigh=ifelse(cihigh>max_dnds_value,max_dnds_value,cihigh),
         VAF_group=factor(VAF_group,levels=VAF_groups$labels[2:length(boundaries)]))%>%
  ggplot(aes(x=VAF_group, y = mle, ymin = cilow, ymax = cihigh, color = name, shape = name)) +
  geom_linerange(position= position_dodge2(width=0.5), size = 0.5, color="darkgrey") +
  geom_point(position=position_dodge2(width=0.5), size = 1.5) +
  labs(y = "Mitochondrial genome-wide dN/dS", x = "Variant allele fraction") +
  theme_classic() + 
  ylim(c(0,max_dnds_value))+
  scale_color_manual(values = mut_type_cols, name="Mutation type") +
  scale_shape_manual(values = c(15, 16, 17, 18), name = "Mutation type") + 
  geom_hline(yintercept=1, linetype='dashed', col = 'darkgrey', size = 0.5) +
  theme_classic()+
  my_theme+
  facet_grid(cols=vars(name))+
  theme(legend.position="none",axis.text.x=element_text(angle=90))

ggsave(filename = paste0(rebuttal_figs_dir,"global_dnds_bytype_by_vaf_plot.pdf"),global_dnds_bytype_by_vaf_plot,width=2.5,height=2.5)

max_dnds_value<-1.5
stats_to_include=c("Missense")
global_dnds_missense_by_vaf_plot<-globaldnds_res_by_vaf%>%
  filter(VAF_group%in%VAF_groups$labels[2:length(boundaries)] & name%in%stats_to_include)%>%
  mutate(cihigh=ifelse(cihigh>max_dnds_value,max_dnds_value,cihigh),
         VAF_group=factor(VAF_group,levels=VAF_groups$labels[2:length(boundaries)]))%>%
  ggplot(aes(x=VAF_group, y = mle, ymin = cilow, ymax = cihigh, color = name, shape = name)) +
  geom_linerange(position= position_dodge2(width=0.75), size = 0.5, color="darkgrey") +
  geom_point(position=position_dodge2(width=0.75), size = 1.5) +
  labs(y = "Mitochondrial genome-wide dN/dS", x = "Variant allele fraction") +
  theme_classic() + 
  ylim(c(0,max_dnds_value))+
  scale_color_manual(values = mut_type_cols, name="Mutation type") +
  scale_shape_manual(values = c(15, 16, 17, 18), name = "Mutation type") + 
  geom_hline(yintercept=1, linetype='dashed', col = 'darkgrey', size = 0.5) +
  theme_classic()+
  my_theme+
  #facet_wrap(~tissue,nrow=1)+
  theme(legend.position="none",axis.text.x=element_text(angle=90))

ggsave(filename = paste0(rebuttal_figs_dir,"global_dnds_missense_by_vaf_plot.pdf"),global_dnds_missense_by_vaf_plot,width=1.6,height=2.5)

```

Now look at a per gene level.
```{r}
#-----------------------------------------------------------------------------------#
##----------------------Get gene-level dNdS info by VAF group---------------
#-----------------------------------------------------------------------------------#
qval_cutoff=0.05
selcv_res_by_vaf<-Map(dndsout=dndscv_by_vaf_blood,VAF_bin=new_VAF_groups,function(dndsout,VAF_bin) {
  temp<-dndsout$sel_cv%>%
    dplyr::select(gene_name,wmis_cv,wnon_cv,qmis_cv,qtrunc_cv)
  
  dplyr::bind_rows(temp%>%dplyr::select(gene_name,"dNdS"=wmis_cv,"qval"=qmis_cv)%>%mutate(type="Missense"),
                   temp%>%dplyr::select(gene_name,"dNdS"=wnon_cv,"qval"=qtrunc_cv)%>%mutate(type="Nonsense"))%>%
        mutate(VAF_group=VAF_bin)
})%>%dplyr::bind_rows()%>%
  mutate(VAF_group=factor(VAF_group,levels=VAF_groups$labels))

dnds_heatmap_by_vaf_by_gene<-selcv_res_by_vaf%>%
  mutate(dnds_if_signif=ifelse(qval<qval_cutoff,paste0(round(dNdS,2),"*"),""),
         gene_name=factor(gene_name,levels=rev(gene_order)))%>%
  filter(VAF_group%in%VAF_groups$labels[2:length(boundaries)])%>%
  dplyr::filter(type=="Missense")%>%
  ggplot(aes(y=gene_name,x=VAF_group,fill=pmin(dNdS,2),label=dnds_if_signif))+
  geom_tile()+
  facet_grid(~type)+
  geom_text(aes(col=type),size=1.6)+
  scale_color_manual(values=c("black","white"),guide="none")+
  theme_classic()+
  my_theme+
  theme(axis.text.x=element_text(angle=90))+
  scale_fill_gradientn(colors=RColorBrewer::brewer.pal(n=11,name = "RdYlGn"),limits=c(0,2))+
  labs(x="Variant allele fraction",y="",fill="dN/dS")

ggsave(filename = paste0(rebuttal_figs_dir,"dnds_heatmap_by_vaf_by_gene.pdf"),dnds_heatmap_by_vaf_by_gene,width=2.5,height=2.5)

```

### dndscv - dichotomous heteroplasmy split
Broadly speaking, the threshold between positive and negative selection is ~20%.  Therefore, to maximise power to detect gene-level effects, run analysis with single cut-off at this threshold.
```{r}

boundaries<-c(0,0.01,0.2,1)
new_VAF_groups=paste0(100*head(boundaries,-1),"-",100*tail(boundaries,-1),"%")
new_VAF_groups=c("<1%","< 20%","> 20%")
VAF_groups=data.frame(labels=new_VAF_groups,LL=head(boundaries,-1),UL=tail(boundaries,-1))

#Assign each observed VAF into its 'VAF group' bin
df_tidy$VAF_group=sapply(df_tidy$vaf,function(vaf) {
  if(vaf==0) {
    return("Absent")
  } else {
    return(VAF_groups$labels[VAF_groups$LL<vaf & VAF_groups$UL>=vaf])
  }
})

dndscv_by_dichotomous_vaf_blood<-lapply(new_VAF_groups,function(VAF_bin) {
  cat(VAF_bin,sep="\n")
  tissue_vaf_info<-df_tidy%>%
    dplyr::filter(VAF_group==VAF_bin)%>%
    separate("mut_ref", c("chr", "pos", "ref", "mut"), "_")%>%
    mutate(pos=as.numeric(pos))%>%
    dplyr::select("sampleID"=exp_ID,chr,pos,ref,mut)%>% 
    dplyr::filter(!duplicated(.))%>% #only count each mutation once per individual in each VAF group
    arrange(sampleID,pos)
  
  mtdna.dndsout <- dndscv(tissue_vaf_info, gene_list=target_genes, 
                          refdb = mtref_rda_path,
                          max_coding_muts_per_sample = Inf,
                          numcode = 2,
                          max_muts_per_gene_per_sample = Inf)
  return(mtdna.dndsout)
})

selcv_res_by_dichotomous_vaf<-Map(dndsout=dndscv_by_dichotomous_vaf_blood,VAF_bin=new_VAF_groups,function(dndsout,VAF_bin) {
  temp<-dndsout$sel_cv%>%
    dplyr::select(gene_name,wmis_cv,wnon_cv,qmis_cv,qtrunc_cv)
  
  dplyr::bind_rows(temp%>%dplyr::select(gene_name,"dNdS"=wmis_cv,"qval"=qmis_cv)%>%mutate(type="Missense"),
                   temp%>%dplyr::select(gene_name,"dNdS"=wnon_cv,"qval"=qtrunc_cv)%>%mutate(type="Nonsense"))%>%
    mutate(VAF_group=VAF_bin)
})%>%dplyr::bind_rows()%>%
  mutate(VAF_group=factor(VAF_group,levels=VAF_groups$labels))

dnds_heatmap_by_dichotomous_vaf_by_gene<-selcv_res_by_dichotomous_vaf%>%
  mutate(dnds_if_signif=ifelse(qval<qval_cutoff,paste0(round(dNdS,2),"*"),""),
         gene_name=factor(gene_name,levels=rev(gene_order)))%>%
  filter(VAF_group%in%VAF_groups$labels[2:length(boundaries)])%>%
  dplyr::filter(type=="Missense")%>%
  ggplot(aes(y=gene_name,x=VAF_group,fill=pmin(dNdS,2),label=dnds_if_signif))+
  geom_tile()+
  facet_grid(~type)+
  geom_text(aes(col=type),size=1.6)+
  scale_color_manual(values=c("black","white"),guide="none")+
  theme_classic()+
  my_theme+
  #theme(axis.text.x=element_text(angle=90))+
  scale_fill_gradientn(colors=RColorBrewer::brewer.pal(n=11,name = "RdYlGn"),limits=c(0,2))+
  labs(x="Variant allele fraction",y="",fill="dN/dS")

ggsave(filename = paste0(rebuttal_figs_dir,"dnds_heatmap_by_dichotomous_vaf_by_gene.pdf"),dnds_heatmap_by_dichotomous_vaf_by_gene,width=2.2,height=2.5)

```

